---
title: "WT_movement_metrics"
author: "Oceane BEGASSAT"
date: "2026-05-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Packages


```{r}
#AKDE
library(tidyverse)
library(ctmm)
library(dplyr)
library(sf)

#Distances
library(tmap)
```

# 1) Home-range size : AKDEc

According to 
https://ecoisilva.github.io/AKDE_minireview/code/AKDE_R-tutorial.html

Vocabulary http://vocab.nerc.ac.uk/collection/MVB/current/.


*AKDE* Tracking data is autocorrelated	
Estimates range distributions from autocorrelated data, by conditioning on an autocorrelation model.

#### Model selection

```{r}
# =========================
# 1. LOAD DATA
# =========================

data <- readRDS("rawdataGPSWT.rds.rds") 


# =========================
# 2. PREP FUNCTION
# =========================


prep_data <- function(df, pop_name, stage_name) {
  df %>%
    filter(pop == pop_name, stage == stage_name) %>%
    dplyr::select(device_id, UTC_datetime, Longitude, Latitude) %>%
    rename(
      ID = device_id,
      timestamp = UTC_datetime,
      longitude = Longitude,
      latitude = Latitude
    )
}

da_incB  <- prep_data(data, "B",  "incub")
da_reaB  <- prep_data(data, "B",  "rear")
da_incGL <- prep_data(data, "GL", "incub")
da_reaGL <- prep_data(data, "GL", "rear")


# =========================
# 3. TELEMETRY
# =========================

tele_incB  <- as.telemetry(da_incB)
tele_reaB  <- as.telemetry(da_reaB)
tele_incGL <- as.telemetry(da_incGL)
tele_reaGL <- as.telemetry(da_reaGL)


# =========================
# 4. MODEL SELECTION FUNCTION
# =========================

# Movement models were selected individually using ctmm.select, starting from an initial guess and using AICc for model comparison.

run_model_selection <- function(tele_list, prefix) {
  
  results <- list()
  
  for (i in seq_along(tele_list)) {
    
    id <- names(tele_list)[i]
    cat("Running:", id, "\n")
    
    dat <- tele_list[[i]]
    
    # Initial guess
    guess <- ctmm.guess(dat, interactive = FALSE)
    
    # Model selection
    fit <- ctmm.select(dat, guess, method = "pHREML", verbose = FALSE)
    
    if (inherits(fit, "ctmm")) {
      
      # Save model
      saveRDS(fit, paste0("NEWmodelwithlocnest_", prefix, "_", id, ".rds"))
      
      # Store results
      results[[i]] <- data.frame(
        ID         = id,
        dataset    = prefix,
        model      = summary(fit)$name,
        dynamics   = fit$dynamics,
        anisotropic = fit$isotropic["sigma"],
        AICc       = fit$AICc,
        logLik     = fit$loglike
      )
      
    } else {
      cat("No valid model for", id, "\n")
    }
  }
  
  # Bind results
  results_df <- bind_rows(results)
  
  # Save table
  write.csv(results_df,
            paste0("model_selectionwithlocnest_", prefix, ".csv"),
            row.names = FALSE)
  
  return(results_df)
}


# =========================
# 5. RUN
# =========================

res_incB  <- run_model_selection(tele_incB,  "incB")
res_reaB  <- run_model_selection(tele_reaB,  "reaB")
res_incGL <- run_model_selection(tele_incGL, "incGL")
res_reaGL <- run_model_selection(tele_reaGL, "reaGL")


# =========================
# 6. FINAL TABLE
# =========================

all_results <- bind_rows(res_incB, res_reaB, res_incGL, res_reaGL)

write.csv(all_results,
          "NEWmodel_selection_ALLwithlocnest.csv",
          row.names = FALSE)
```

#### Apply model

```{r}

# ===========================
#  Load Data
# ===========================
input_dir  <- "~/guif2024_akde/data2"    # where the selected models are stored
output_dir <- "~/guif2024_akde/results2" # where to store the results
data_file  <- "~/guif2024_akde/dataNEW/rawdataGPSWT.rds.rds" # dataset

data1 <- readRDS(data_file)

# =========================
# 2. PREP FUNCTION
# =========================
prep_data <- function(df, pop_name, stage_name){
  
  df %>%
    filter(pop == pop_name, stage == stage_name) %>%
    dplyr::select(device_id, UTC_datetime, Longitude, Latitude) %>%
    rename(
      ID = device_id,
      timestamp = UTC_datetime,
      longitude = Longitude,
      latitude = Latitude
    )
}

# datasets
da_incB  <- prep_data(data1, "B",  "incub")
da_reaB  <- prep_data(data1, "B",  "rear")
da_incGL <- prep_data(data1, "GL", "incub")
da_reaGL <- prep_data(data1, "GL", "rear")

# =========================
# 3. TELEMETRY
# =========================
tele_incB  <- as.telemetry(da_incB)
tele_reaB  <- as.telemetry(da_reaB)
tele_incGL <- as.telemetry(da_incGL)
tele_reaGL <- as.telemetry(da_reaGL)

cat("✅ Data processing done : telemetry data are loaded\n")

tele_list <- list(
  incB  = tele_incB,
  reaB  = tele_reaB,
  incGL = tele_incGL,
  reaGL = tele_reaGL
)

# =========================================
#   Fonction de calcul AKDE pour un individu
# =========================================

#File structure NEWmodelwithlocnest_incB_243695.rds

compute_akde <- function(fit_file, tele_list, weighted=FALSE) {
  fname <- basename(fit_file)
  prefix <- sub("^NEWmodelwithlocnest_", "", sub("\\.rds$", "", fname))   # ex: reaGL_243695
  pop_stage <- sub("_.*$", "", prefix)                      # ex: reaGL
  indiv_name <- sub("^[^_]*_", "", prefix)                  # ex: 243696

  # Load selected model
  fit <- readRDS(fit_file)

  # Extract telemetry object
  ind <- tele_list[[pop_stage]][[indiv_name]]

  #  AKDE calculation
  UD <- akde(ind, fit, weights=weighted)
  ci <- summary(UD)$CI
  N_eff <- summary(UD)$DOF["area"]
  N_abs <- nrow(ind)

  # save
  saveRDS(UD, file.path(output_dir, paste0("NEWUDakdec_", prefix, ".rds")))
  pdf(file.path(output_dir, paste0("NEWUDakdec_", prefix, ".pdf")))
  plot(UD, main=paste("NEWUDakdec -", prefix))
  dev.off()

 #Export 
 ud_sf <- ctmm::as.sf(UD, level.UD=0.95)
 shp_file <- file.path(output_dir, paste0("NEWUDakdec_withlocnest_", prefix, ".shp"))
 sf::st_write(ud_sf, shp_file, delete_layer = TRUE)

  # Summary
  return(data.frame(
    Individual = prefix,
    Areakm2       = ci[2],
    CI_low     = ci[1],
    CI_high    = ci[3],
    N_eff = N_eff,
    N_abs = N_abs,
    stringsAsFactors = FALSE
  ))
}

# ===========================
#   Application à tous les modèles
# ===========================

model_files <- list.files(
  input_dir,
  pattern = "^NEWmodelwithlocnest_.*\\.rds$",
  full.names = TRUE
)

cat("Models selected:\n")
print(basename(model_files))

results_allakdec <- do.call(
  rbind,
  lapply(model_files, compute_akde, tele_list=tele_list, weighted=FALSE)
)

# Save
saveRDS(results_allakdec, file.path(output_dir, "NEWakdec_results_withlocnest.rds"))
write.csv(results_allakdec, file.path(output_dir,"NEWakdec_results_withlocnest.csv"), row.names=FALSE)

cat("✅ AKDE  for" ,nrow(results_allakdec), "individus\n")
```

# 2) Mean daily distance to the nest

### Load data

```{r}
datal <- readRDS("rawdataGPSWT.rds" )#one row per location, n = 239 897

#number of days in each breeding period per individual
d_days <- datal %>%
  mutate(
    days_incub = as.numeric(days_incub),
    days_rear  = as.numeric(days_rear),
    tot_days   = as.numeric(tot_days)
  ) %>%
  group_by(device_id) %>%
  summarise(
    nb_locs      = n(),
    nb_locs_inc  = sum(stage == "incub", na.rm = TRUE),
    nb_locs_rea  = sum(stage == "rear",  na.rm = TRUE),
    start_inc_date = min(date[stage == "incub"], na.rm = TRUE),
    end_inc_date = max(date[stage == "incub"], na.rm = TRUE),
    days_incub   = first(days_incub),
    days_rear    = first(days_rear),
    tot_days     = first(tot_days),
    .groups = "drop"
  )


datal <- datal %>%
  mutate(across(c(device_id, bague, pop, colonie, sex, name, stage), as.factor))
datal <- datal %>% select(device_id, name, bague, pop, colonie, sex, stage, UTC_datetime, date, X, Y, Latitude, Longitude, LON_nid_R, LAT_nid_R, X_nid, Y_nid)
summary(datal)

```

### Distance to the nest
```{r}
#Individual locations; split by study area
dataB <- datal %>% 
  filter(pop == "B") #X and Y ar UTM 30 and 31N coordinates
datagl <- datal %>%  
  filter(pop == "GL")

#nests coordinates as UTM
nestB <- datal %>% 
  filter(pop == "B") %>% 
  group_by(device_id) %>% 
  summarize(X_nid = first(X_nid), Y_nid = first(Y_nid))

nestGL <- datal %>% 
  filter(pop == "GL") %>% 
  group_by(device_id) %>% 
  summarize(X_nid = first(X_nid), Y_nid = first(Y_nid))

#convert as spatial: sf
nestB <- nestB %>% 
  st_as_sf(coords = c("X_nid", "Y_nid"), crs = 32631)  # Zone 31N pour B
dataB <- dataB %>%  st_as_sf(coords = c("X", "Y"), crs = 32631)

nestGL <- nestGL %>% 
  st_as_sf(coords = c("X_nid", "Y_nid"), crs = 32630)
datagl <- datagl %>%  st_as_sf(coords = c("X", "Y"), crs = 32630)

#maps to check
tmap_mode("view")  

trajB <- dataB %>%
  group_by(device_id) %>%
  arrange(device_id, UTC_datetime) %>% 
  summarize(do_union = FALSE) %>%
  st_cast("LINESTRING")

trajGL <- datagl %>%
  group_by(device_id) %>%
  arrange(device_id, UTC_datetime) %>%
  summarize(do_union = FALSE) %>%
  st_cast("LINESTRING")

tm_shape(trajB) +
  tm_lines(col = "device_id", palette = "Set1", lwd = 0.5, group = "Trajectoires B") +
tm_shape(nestB) +
  tm_symbols(col = "device_id", shape = 21, size = 0.0001, alpha = 1, group = "Nids B") +

tm_shape(trajGL) +
  tm_lines(col = "device_id", palette = "Set2", lwd = 0.5, group = "Trajectoires GL") +
tm_shape(nestGL) +
  tm_symbols(col = "device_id", shape = 21, size = 0.0001, alpha = 1, group = "Nids GL") +

tm_layout(legend.outside = TRUE)


#--------------- Distances to the nest ----------------

#Distance between each loc and the individual's nest

dataB$DistNest <- NA  # to store distances
for (i in unique(dataB$device_id)) {
  # nest of the individual
  nest <- nestB %>% filter(device_id == i)
  # nest coords
  nest_coords <- st_coordinates(nest)
  # Locs for this individual
  individual_locs <- dataB %>% filter(device_id == i)
  # Distances calculation
  individual_locs$DistNest <- as.numeric(st_distance(individual_locs, nest)) / 1000 # km

  dataB$DistNest[dataB$device_id == i] <- individual_locs$DistNest
}
summary(dataB$DistNest) 
 

datagl$DistNest <- NA  
for (i in unique(datagl$device_id)) {
  nest <- nestGL %>% filter(device_id == i)
  
  nest_coords <- st_coordinates(nest)
  
  individual_locs <- datagl %>% filter(device_id == i)
  
  individual_locs$DistNest <- as.numeric(st_distance(individual_locs, nest)) / 1000  
  
  datagl$DistNest[datagl$device_id == i] <- individual_locs$DistNest
}
summary(datagl$DistNest) 

#store coordinates
dataB <- dataB %>% 
  mutate(X = st_coordinates(.)[,1],
         Y = st_coordinates(.)[,2])

datagl <- datagl %>% 
  mutate(X = st_coordinates(.)[,1],
         Y = st_coordinates(.)[,2])

dataB <- dataB %>% st_drop_geometry()
datagl <- datagl %>% st_drop_geometry()

datall <- bind_rows(dataB, datagl)

```

### Remove locations on the nest and summarise

```{r}
# Add a column to indicate if the location is within 12m of the nest

datall <- datall %>%
  mutate(near_nest = ifelse(DistNest <= 0.012, 1, 0))  # 1 = near the nest, 0 = away


datall1 <- datall %>% 
  filter(DistNest>0.012) #114263 rows


quantile(datall1$DistNest, 0.95)

datall1B <- datall1 %>% filter(pop == "B")
quantile(datall1B$DistNest, 0.95)

datall1GL <- datall1 %>% filter(pop == "GL")
quantile(datall1GL$DistNest, 0.95)

#summary per day, distnest : min, med, max, mean, sd, 95IC per DAY and individual
stats_dist_day <- datall1 %>%
  group_by(device_id, date) %>%
  summarise(
    n = n(),
    stage = first(stage), 
    sex = first(sex), 
    pop = first(pop), 
    min_dist = min(DistNest, na.rm = TRUE),
    median_dist = median(DistNest, na.rm = TRUE),
    max_dist = max(DistNest, na.rm = TRUE),
    p95_dist = quantile(DistNest, 0.95, na.rm = TRUE),
    mean_dist = mean(DistNest, na.rm = TRUE),
    sd_dist = sd(DistNest, na.rm = TRUE),
    se = sd_dist / sqrt(n),
    ic95_low = mean_dist - 1.96 * se,
    ic95_up = mean_dist + 1.96 * se,
    .groups = "drop"
  )

```

# 3) Movement intensity

### Cumulative distances per day
```{r}

# 1.Separate by pop
dataB <- datall1B 
datagl  <- datall1GL 

# 2. Convert as sf
dataB_sf <- st_as_sf(dataB, coords = c("X", "Y"), crs = 32631)  # UTM zone 31N
datagl_sf  <- st_as_sf(datagl, coords = c("X", "Y"), crs = 32630)   # UTM zone 30N

# 3. Function to calculate cumulative distances
calcul_distance_par_jour <- function(df_sf) {
  df_sorted <- df_sf %>%
    group_by(device_id, date) %>% #per day
    arrange(UTC_datetime) %>% 
    mutate( 
      prev_geom = lag(geometry),  # geometry of the previous location
      dist_step = st_distance(prev_geom, geometry, by_element = TRUE),#distance (m)
      dist_step = as.numeric(dist_step)
    ) %>% #sum across the day
    summarise(
      dist_totale_jour = sum(dist_step, na.rm = TRUE),
      pop = first(pop),
      stage = first(stage), 
      sex = first(sex), 
      n = n(), #nb locs
      .groups = "drop"
    )
  
  return(df_sorted)
}

# 4. Appliquer la fonction aux deux jeux
dist_brenne <- calcul_distance_par_jour(dataB_sf)
summary(dist_brenne)
 
dist_glieu  <- calcul_distance_par_jour(datagl_sf)
summary(dist_glieu)

# add to a pooled dataset
dist_brenne <- st_drop_geometry(dist_brenne)
dist_glieu <- st_drop_geometry(dist_glieu)
dist_all <- bind_rows(dist_brenne, dist_glieu)


#effects of pop and sex per stage
ggplot(dist_all %>% filter(stage == "incub"), aes(x = pop, y = dist_totale_jour, fill = sex)) +
  geom_boxplot(position = position_dodge(0.8)) +
  labs(title = "Daily cumulative distances per sex and population (Incubation)",
       x = "Population", y = " (m)") +
  scale_fill_brewer(palette = "Set1") +
  theme_classic()

ggplot(dist_all %>% filter(stage == "rear"), aes(x = pop, y = dist_totale_jour, fill = sex)) +
  geom_boxplot(position = position_dodge(0.8)) +
  labs(title = "Daily cumulative distances per sex and population (Rearing)",
       x = "Population", y = " (m)") +
  scale_fill_brewer(palette = "Set1") +
  theme_classic()


```


### Divide by the sum of delta t
Average movement intensity

```{r}

# 3. Function to calculate delta t
calcul_deltat <- function(df) {

  df %>%
    arrange(device_id, date, UTC_datetime) %>%
    group_by(device_id, date) %>%# per day
    mutate(
      # time between locs (sec)
      deltat = as.numeric(difftime(
        UTC_datetime,
        lag(UTC_datetime),
        units = "secs"
      )),

      # distance entre points (en mètres)
      dist_step = as.numeric(st_distance(
        geometry,
        lag(geometry),
        by_element = TRUE
      ))
    ) %>%
    summarise(
      dist_totale_jour = sum(dist_step, na.rm = TRUE),   # m
      temps_total_jour = sum(deltat, na.rm = TRUE),      # s
      vitesse_moy_jour = dist_totale_jour / temps_total_jour, # m/s
      vitesse_kmh = (dist_totale_jour / temps_total_jour) * 3.6,
      pop = first(pop),
      stage = first(stage),
      sex = first(sex),
      n=n(),#locs par jour
      .groups = "drop"
    )
}

# 4. Apply
vitmoyB <- calcul_deltat(dataB_sf)
vitmoygl <- calcul_deltat(datagl_sf)
summary(vitmoyB$vitesse_kmh) #0.59-10.9km/h 
summary(vitmoygl$vitesse_kmh) #1.7-8.34 km/h

vitmoyB <- st_drop_geometry(vitmoyB)
vitmoygl <- st_drop_geometry(vitmoygl)

# Merge
vitall <- bind_rows(vitmoyB, vitmoygl)
hist(vitall$vitesse_kmh) #right squewed
summary(vitall$vitesse_kmh)

ggplot(vitall, aes(x = stage, y = vitesse_kmh, fill = sex)) +
  geom_boxplot(position = position_dodge(0.8)) +
  labs(title = "",
       x = "stage", y = " (m)") +
  scale_fill_brewer(palette = "Set1") +
  theme_classic()

ggplot(vitall %>% filter(pop == "GL"), aes(x = stage, y = vitesse_kmh, fill = sex)) +
  geom_boxplot(position = position_dodge(0.8)) +
  labs(title = "",
       x = "stage", y = " (m)") +
  scale_fill_brewer(palette = "Set1") +
  theme_classic()

ggplot(vitall %>% filter(pop == "B"), aes(x = stage, y = vitesse_kmh, fill = sex)) +
  geom_boxplot(position = position_dodge(0.8)) +
  labs(title = "",
       x = "stage", y = " (m)") +
  scale_fill_brewer(palette = "Set1") +
  theme_classic()

ggplot(vitall, aes(x = stage, y = vitesse_kmh, fill = pop)) +
  geom_boxplot(position = position_dodge(0.8)) +
  labs(title = "",
       x = "stage", y = " (m)") +
  scale_fill_brewer(palette = "Set1") +
  theme_classic()

```

