---
title: "Behav_classif_MomentuHMM"
author: "Oceane BEGASSAT"
date: "2026-05-04"
output: html_document
---


#Packages
```{r, warning = FALSE, echo = FALSE}
library(tidyverse)
library(momentuHMM)
library(sf)
library(amt)
library(zoo)  
library(data.table)
library(gridExtra)
library(patchwork) #panel of plots
library(tmap)
```

# Data processing

##Load data
```{r}
data_day <- readRDS("dataHMM.rds")
```

##1min resampling
```{r}
#From data_day, apply a filter (tolerance = 5sec) and reasmple to the same interval

data_day_lowdelta <- data_day %>% 
  group_by(device_id_date) %>% 
  arrange(UTC_datetime) %>% 
  mutate(time_diff = difftime(UTC_datetime, lag(UTC_datetime), units = "secs")) %>%
  filter(time_diff < 65 & time_diff >55)

#Generate all possible series of regular timestamps (every 60 seconds) by dragging the starting point to the first 15 locations.
# For each attempt:
# Resample the data according to this starting point.
# Count the number of points actually associated with real timestamps (with a tolerance of ±5 seconds).
# Keep the version that maximises the number of points recovered

resample_60s_tolerance <- function(df, step = 60, tolerance = 5, start_index = 1) {
  df <- df %>% arrange(UTC_datetime)

  # We start from the nth observation to construct the grid.
  start_time <- df$UTC_datetime[start_index]
  end_time <- max(df$UTC_datetime)

# Expected sequence (step = 60s)
  time_seq <- seq(from = start_time, to = end_time, by = step)
  
  #For each theoretical point, we look for the closest actual point within ±5s.
  matched_points <- purrr::map_dfr(time_seq, function(t_target) {
    window <- df %>%
      filter(abs(as.numeric(difftime(UTC_datetime, t_target, units = "secs"))) <= tolerance)
    
    if (nrow(window) == 0) return(NULL)  # Si pas de point dans la fenêtre
    
    # keep it
    closest_row <- window %>%
      slice_min(abs(as.numeric(difftime(UTC_datetime, t_target, units = "secs")))) %>%
      mutate(resampled_time = t_target)
    
    return(closest_row)
  })
  
  return(matched_points)
}

resample1nona <- data_day_lowdelta %>%
  group_by(device_id_date) %>%
  group_split() %>%
  map_dfr(~resample_60s_tolerance(.x, tolerance = 5, start_index = 1)) 

resample1nona %>%
  group_by(device_id_date) %>%
  summarise(n = n()) %>%
  summary()

resample2nona <- data_day_lowdelta %>%
  group_by(device_id_date) %>%
  group_split() %>%
  map_dfr(~resample_60s_tolerance(.x, tolerance = 5, start_index = 2)) 

resample3nona <- data_day_lowdelta %>%
  group_by(device_id_date) %>%
  group_split() %>%
  map_dfr(~resample_60s_tolerance(.x, tolerance = 5, start_index = 3)) 

resample4nona <- data_day_lowdelta %>%
  group_by(device_id_date) %>%
  group_split() %>%
  map_dfr(~resample_60s_tolerance(.x, tolerance = 5, start_index = 4)) #--> best one

resample5nona <- data_day_lowdelta %>%
  group_by(device_id_date) %>%
  group_split() %>%
  map_dfr(~resample_60s_tolerance(.x, tolerance = 5, start_index = 5))

resample4bisnona <- data_day %>%
  group_by(device_id_date) %>%
  group_split() %>%
  map_dfr(~resample_60s_tolerance(.x, tolerance = 5, start_index = 4))


#keep resample4 that maximises the data number
data_1minnona <- resample4nona

#By removing rows where no data was found to match the timestep we can easily have a look on the gaps between consective locations.

#check
# time diff between consecutive locs
time_diffs <- data_1minnona %>%
  group_by(device_id_date) %>%
  arrange(UTC_datetime) %>%
  mutate(delta_sec = as.numeric(difftime(UTC_datetime, lag(UTC_datetime), units = "secs"))) %>%
  ungroup()

plot(time_diffs$delta_sec)
hist(time_diffs$delta_sec)
summary(time_diffs$delta_sec) 


#Keep the theoretical time grid with NA if no match is found

resample_60s_tolerance <- function(df, step = 60, tolerance = 5, start_index = 1) {
  df <- df %>% arrange(UTC_datetime)

  start_time <- df$UTC_datetime[start_index]
  end_time <- max(df$UTC_datetime)

  time_seq <- seq(from = start_time, to = end_time, by = step)

  # Colonnes statiques à préserver même en cas de NA
  static_cols <- c("name", "device_id", "sex", "pop", "device_id_date")
  static_values <- df[start_index, static_cols]

  matched_points <- purrr::map_dfr(time_seq, function(t_target) {
    window <- df %>%
      filter(abs(as.numeric(difftime(UTC_datetime, t_target, units = "secs"))) <= tolerance)

    if (nrow(window) == 0) {
      # Ligne vide avec NA + valeurs statiques
      empty_row <- tibble(resampled_time = t_target)

      # Ajouter les colonnes dynamiques (tout sauf static_cols et UTC_datetime)
      dynamic_cols <- setdiff(names(df), c(static_cols, "UTC_datetime"))
      for (col in dynamic_cols) {
        empty_row[[col]] <- NA
      }

      # Ajouter les colonnes statiques avec leurs valeurs
      for (col in static_cols) {
        empty_row[[col]] <- static_values[[col]]
      }

      # Ajouter la colonne UTC_datetime à NA
      empty_row$UTC_datetime <- NA

      return(empty_row)
    }

    closest_row <- window %>%
      slice_min(abs(as.numeric(difftime(UTC_datetime, t_target, units = "secs")))) %>%
      mutate(resampled_time = t_target)

    return(closest_row)
  })

  return(matched_points)
}

resample1 <- data_day_lowdelta %>%
  group_by(device_id_date) %>%
  group_split() %>%
  map_dfr(~resample_60s_tolerance(.x, tolerance = 5, start_index = 1))

resample2 <- data_day_lowdelta %>%
  group_by(device_id_date) %>%
  group_split() %>%
  map_dfr(~resample_60s_tolerance(.x, tolerance = 5, start_index = 2))

resample3 <- data_day_lowdelta %>%
  group_by(device_id_date) %>%
  group_split() %>%
  map_dfr(~resample_60s_tolerance(.x, tolerance = 5, start_index = 3))

resample4 <- data_day_lowdelta %>%
  group_by(device_id_date) %>%
  group_split() %>%
  map_dfr(~resample_60s_tolerance(.x, tolerance = 5, start_index = 4))

resample5 <- data_day_lowdelta %>%
  group_by(device_id_date) %>%
  group_split() %>%
  map_dfr(~resample_60s_tolerance(.x, tolerance = 5, start_index = 5))

resample4bis <- data_day %>%
  group_by(device_id_date) %>%
  group_split() %>%
  map_dfr(~resample_60s_tolerance(.x, tolerance = 5, start_index = 4))

#how many locs whithout NA
resample1 %>%
  filter(!is.na(X) & !is.na(Y)) %>%
  summarise(n_valid = n()) 

resample2 %>%
  filter(!is.na(X) & !is.na(Y)) %>%
  summarise(n_valid = n())

resample3 %>%
  filter(!is.na(X) & !is.na(Y)) %>%
  summarise(n_valid = n())

resample4 %>%
  filter(!is.na(X) & !is.na(Y)) %>%
  summarise(n_valid = n())

resample5 %>%
  filter(!is.na(X) & !is.na(Y)) %>%
  summarise(n_valid = n())

resample4bis %>%
  filter(!is.na(X) & !is.na(Y)) %>%
  summarise(n_valid = n())

sum(is.na(resample4$X))

data_1min <- resample4

```


## 1min Non linear interpolation in gaps = 2min (keep NA in others)


More "realistic" interpolation method and/or account for the location uncertainty during the gaps. 
This is what the crawlWrap (complicated : several parameters) and MIfitHMM functions are for.

CRAWL intègre une modélisation réaliste des trajectoires et la prise en compte de l’erreur de mesure.
Nécessite coordonnées projetées UTM
2 different UTM zones


```{r}
#interpolate when there is only one consecutive NA
#Split by pop
data1minB <- data_1min %>% 
  filter(pop == "B")

data1minGL <- data_1min %>% 
  filter(pop == "GL")

## FUNCTIONS ##

interpolate_small_gaps <- function(df) {
  df <- df %>% arrange(resampled_time)

  df <- df %>%
    mutate(
      valid = !(is.na(X) | is.na(Y)),
      prev_valid = lag(valid),
      next_valid = lead(valid),
      is_single_na = !valid & prev_valid & next_valid
    )

  na_idx <- which(df$is_single_na)
  if (length(na_idx) == 0) return(df)

  # minimal df for crawl
  df_crawl_input <- df %>%
    select(device_id_date, resampled_time, X, Y) %>%
    rename(ID = device_id_date, time = resampled_time)

  # use of crawlWrap
  crawl_result <- tryCatch({
    crawlWrap(
      obsData = df_crawl_input,
      timeStep = "1 min",
      coord = c("X", "Y"),
      time.name = "time"
    )
  }, error = function(e) {
    return(NULL)
  })

  # checking
  if (is.null(crawl_result) || !"crwPredict" %in% names(crawl_result)) return(df)

  predicted <- crawl_result$crwPredict

  # Replace isolated NAs with predicted coordinates
  df$X[na_idx] <- predicted$mu.x[na_idx]
  df$Y[na_idx] <- predicted$mu.y[na_idx]

  return(df)
}

can_run_crawl <- function(df) {
  df <- df %>% arrange(resampled_time)

  # At least 3 valid points + at least one NA
  enough_points <- df %>% filter(!is.na(X) & !is.na(Y)) %>% nrow() >= 3
  has_na <- any(is.na(df$X) | is.na(df$Y))

  return(enough_points & has_na)
}

########

data1min_interpoB <- data1minB %>%
  group_by(device_id_date) %>%
  group_split() %>%
  map_dfr(function(df) {
    if (can_run_crawl(df)) {
      interpolate_small_gaps(df)  
    } else {
      df  
    }
  })

data1min_interpoGL <- data1minGL %>%
  group_by(device_id_date) %>%
  group_split() %>%
  map_dfr(function(df) {
    if (can_run_crawl(df)) {
      interpolate_small_gaps(df)  
    } else {
      df  
    }
  })


colSums(is.na(data1min_interpoB))
colSums(is.na(data1min_interpoGL))
```

##1min Check interpolations

Check if interpolations are reliable
```{r}
#Spatial plots to see if interpolations are reliable
data1min_interpoB %>%
  filter(valid | is_single_na) %>%
  group_by(device_id_date) %>%
  group_split() %>%
  walk(function(df) {
    gg <- ggplot(df %>% arrange(resampled_time)) +
      geom_path(aes(x = X, y = Y), color = "grey70") +
      geom_point(aes(x = X, y = Y, color = is_single_na)) +
      scale_color_manual(values = c("FALSE" = "blue", "TRUE" = "red"),
                         labels = c("Valide", "Interpolé"),
                         name = "Type de point") +
      coord_equal() +
      labs(
        title = paste("Trajectoire :", unique(df$device_id_date)),
        x = "X", y = "Y"
      ) +
      theme_minimal()
    
    print(gg)
    Sys.sleep(1)
  })

#Spatial plots to see if interpolations are reliable
data1min_interpoGL %>%
  filter(valid | is_single_na) %>%
  group_by(device_id_date) %>%
  group_split() %>%
  walk(function(df) {
    gg <- ggplot(df %>% arrange(resampled_time)) +
      geom_path(aes(x = X, y = Y), color = "grey70") +
      geom_point(aes(x = X, y = Y, color = is_single_na)) +
      scale_color_manual(values = c("FALSE" = "blue", "TRUE" = "red"),
                         labels = c("Valide", "Interpolé"),
                         name = "Type de point") +
      coord_equal() +
      labs(
        title = paste("Trajectoire :", unique(df$device_id_date)),
        x = "X", y = "Y"
      ) +
      theme_minimal()
    
    print(gg)
    Sys.sleep(1)
  })

```

## 1min Give an ID to each sequence
```{r}

assign_sequence_ids_no_na <- function(df) {
  df <- df %>% arrange(resampled_time)

  # Identify whether each line is valid (X or Y not NA)
  df <- df %>%
    mutate(is_valid = !(is.na(X) | is.na(Y)))

  # Run Length Encoding 
  rle_valid <- rle(df$is_valid)
  lengths <- rle_valid$lengths
  values <- rle_valid$values

  group_ids <- rep(NA_integer_, nrow(df))
  index <- 1
  group_num <- 1

  for (i in seq_along(lengths)) {
    len <- lengths[i]
    val <- values[i]

    if (val) {
      # Sequence is fully valid → it is assigned an ID
      group_ids[index:(index + len - 1)] <- group_num
      group_num <- group_num + 1
    }
    # If it is not valid (NA in X or Y), no ID is given.
    index <- index + len
  }

  df <- df %>%
    mutate(device_id_date_num = ifelse(is.na(group_ids), NA, paste0(device_id_date, "_", group_ids)))

  return(df)
}

data1min_interpocutB <- data1min_interpoB %>%
  group_by(device_id_date) %>%
  group_split() %>%
  map_dfr(assign_sequence_ids_no_na)

data1min_interpocutGL <- data1min_interpoGL %>%
  group_by(device_id_date) %>%
  group_split() %>%
  map_dfr(assign_sequence_ids_no_na)

data1min_interpocutGL$device_id_date_num <- as.factor(data1min_interpocutGL$device_id_date_num)
data1min_interpocutB$device_id_date_num <- as.factor(data1min_interpocutB$device_id_date_num)

mean(is.na(data1min_interpocutB$device_id_date_num))  
mean(is.na(data1min_interpocutGL$device_id_date_num)) 

size_seqence <- data1min_interpocutB %>%
  group_by(device_id_date_num) %>%
  summarise(n = n(), .groups = "drop")

size_seqence <- data1min_interpocutGL %>%
  group_by(device_id_date_num) %>%
  summarise(n = n(), .groups = "drop")

#n = 10 data per sequence at least

seq1min10GL <- data1min_interpocutGL %>%
  filter(!device_id_date_num == "NA") %>%
  group_by(device_id_date_num) %>%
  filter(n() >= 10)
seq1min10B <- data1min_interpocutB %>%
  filter(!device_id_date_num == "NA") %>%
  group_by(device_id_date_num) %>%
  filter(n() >= 10)

#No NA in X et Y
colSums(is.na(seq1min10B))
colSums(is.na(seq1min10GL))

```

## 1min convert XY to LL

```{r}
# Convert the new coordinates (interpolated or not) X and Y to Lon Lat for the next step.

# Transform to longitude/latitude (EPSG:4326)
# seq1min10Bll <- st_transform(seq1min10Bsf, crs = 4326)
# seq1min10GLll <- st_transform(seq1min10GLsf, crs = 4326)
# seq1min15Bll <- st_transform(seq1min15Bsf, crs = 4326)
# seq1min15GLll <- st_transform(seq1min15GLsf, crs = 4326)
# 
# # Extract coordinates
# coords <- st_coordinates(seq1min10Bll)
# seq1min10B$lon <- coords[,1]
# seq1min10B$lat <- coords[,2]
# 
# coords <- st_coordinates(seq1min10GLll)
# seq1min10GL$lon <- coords[,1]
# seq1min10GL$lat <- coords[,2]
# 
# colSums(is.na(seq1min10B)) #no na
# colSums(is.na(seq1min10GL))
# 
# #Combine the two pop sets into one df
# #Keep all columns except Latitude and Longitude
# #Remove the Latitude and Longitude columns from both sets to avoid confusion
# seq1min10B_clean <- seq1min10B %>% select(-Latitude, -Longitude)
# seq1min10GL_clean <- seq1min10GL %>% select(-Latitude, -Longitude)
# 
# # Combined 
# seq_all <- bind_rows(seq1min10B_clean, seq1min10GL_clean)#83810 rows
# colSums(is.na(seq_all))
```

# HMM

### Prepare Data
```{r}
#Data with at least 10 consecutive locs per sequence (tested with 15, not shown), and an ID per sequence (consecutive locs) 
data1min_interpocutB <- data1min_interpoB %>%
  group_by(device_id_date) %>%
  group_split() %>%
  map_dfr(assign_sequence_ids_no_na)

data1min_interpocutGL <- data1min_interpoGL %>%
  group_by(device_id_date) %>%
  group_split() %>%
  map_dfr(assign_sequence_ids_no_na)

data1min_interpocutGL$device_id_date_num <- as.factor(data1min_interpocutGL$device_id_date_num)
data1min_interpocutB$device_id_date_num <- as.factor(data1min_interpocutB$device_id_date_num)

#No NA remaining in X and Y
dataforprep1min10GL <- seq1min10GL %>% 
  select(device_id_date_num, resampled_time, X, Y) %>%
  rename(ID = device_id_date_num, time = resampled_time) %>%
  arrange(ID, time) %>% 
  ungroup()

dataforprep1min10B <- seq1min10B %>% 
  select(device_id_date_num, resampled_time, X, Y) %>%
  rename(ID = device_id_date_num, time = resampled_time) %>%
  arrange(ID, time) %>% 
  ungroup()

colSums(is.na(dataforprep1min10GL)) 
colSums(is.na(dataforprep1min10B)) #no NA

dataforprep1min10GL <- as.data.frame(dataforprep1min10GL)
dataforprep1min10B <- as.data.frame(dataforprep1min10B)

#with covariates

#pop, sex
dataforprep1min10covGL <- seq1min10GL %>% 
  select(device_id_date_num, resampled_time, X, Y, sex, pop) %>%
  rename(ID = device_id_date_num, time = resampled_time) %>%
  arrange(ID, time) %>% 
  ungroup()
dataforprep1min10covGL <- as.data.frame(dataforprep1min10covGL)

dataforprep1min10covB <- seq1min10B %>% 
  select(device_id_date_num, resampled_time, X, Y, sex, pop) %>%
  rename(ID = device_id_date_num, time = resampled_time) %>%
  arrange(ID, time) %>% 
  ungroup()
dataforprep1min10covB <- as.data.frame(dataforprep1min10covB)

###

#according to Theo Michelot & Roland Langrock May 8, 2023 (prepData is only used with regular data)
prep1min10GL <- prepData(data = dataforprep1min10GL, type = "UTM", coordNames = c("X", "Y"))
prep1min10B <- prepData(data = dataforprep1min10B, type = "UTM", coordNames = c("X", "Y"))

#autocorrelation : we can check the autocorrelation function to see if our step lengths show cyclical pattern (they don't)
acf(prep1min10GL$step[!is.na(prep1min10GL$step)],lag.max=100)
acf(prep1min10B$step[!is.na(prep1min10B$step)],lag.max=100)
plot(prep1min10GL) #step length 0-500, turning angles  = -pi - pi (for each individual)
summary(prep1min10GL$step) #mean = 73m, max = 1.1km
sd(prep1min10GL$step, na.rm = T) #sd = 112m
summary(prep1min10GL$angle) #mean = 0, med =0
sd(prep1min10GL$angle, na.rm = T) #2.04

plot(prep1min10B)
summary(prep1min10B$step)# mean = 69, max = 0.9km
sd(prep1min10B$step, na.rm = T) #sd = 136.5 m
summary(prep1min10B$angle)
sd(prep1min10B$angle, na.rm = T) #sd = 2.10
#drome tuto Marie Gilbertson 2023-05-04 https://mjones029.github.io/Tutorials/HMM_tutorial.html

```

### Remove NA 
(limit the number of NA even if their occurence is not a problem, limit potential bias)
```{r}
#GL
length(unique(prep1min10GL$ID)) #device_annee_mois_jour_numseq
colSums(is.na(prep1min10GL)) #for the first and las loc of the day
attr(prep1min10GL, "class")
inherits(prep1min10GL, "momentuHMMData")

#NA = first of agroupeID ?
prep1min10GL %>%
  group_by(ID) %>%
  mutate(row = row_number()) %>%
  filter(row == 1 | lag(is.na(step)) | is.na(step)) %>%
  select(ID, row, time, step, angle) %>%
  arrange(ID, row)

# The first in each sequence has NA for the angle.
# And the last in each sequence has NA for step and angle.

### Remove the first and last of each sequence

prep1min10cleanGL <- prep1min10GL %>%
  arrange(ID, time) %>%
  group_by(ID) %>%
  mutate(row = row_number(), last_row = n()) %>%
  filter(row != 1 & row != last_row) %>%
  ungroup() %>%
  select(-row, -last_row)

colSums(is.na(prep1min10cleanGL)) # 39 NA ok

#keep the proper class (and the attributes)
inherits(prep1min10GL, "momentuHMMData")
inherits(prep1min10cleanGL, "momentuHMMData")
class(prep1min10cleanGL) <- c("momentuHMMData", class(prep1min10cleanGL))
inherits(prep1min10cleanGL, "momentuHMMData")


#B do the same steps
length(unique(prep1min10B$ID)) 
colSums(is.na(prep1min10B)) 
attr(prep1min10B, "class")
inherits(prep1min10B, "momentuHMMData")

prep1min10B %>%
  group_by(ID) %>%
  mutate(row = row_number()) %>%
  filter(row == 1 | lag(is.na(step)) | is.na(step)) %>%
  select(ID, row, time, step, angle) %>%
  arrange(ID, row)

prep1min10cleanB <- prep1min10B %>%
  arrange(ID, time) %>%
  group_by(ID) %>%
  mutate(row = row_number(), last_row = n()) %>%
  filter(row != 1 & row != last_row) %>%
  ungroup() %>%
  select(-row, -last_row)

colSums(is.na(prep1min10cleanB)) 
inherits(prep1min10B, "momentuHMMData")
inherits(prep1min10cleanB, "momentuHMMData")
class(prep1min10cleanB) <- c("momentuHMMData", class(prep1min10cleanGL))
inherits(prep1min10cleanB, "momentuHMMData")
```



### Params distribution
```{r}
#GL
#--STEP LENGTH----

# Plot histogram of step lengths 
hist(prep1min10cleanGL$step, xlab = "step length (m)", main = "", breaks = 50, xlim = c(0, 500)) #in meters (UTM)
#asymetric, try in log sclae
hist(log(prep1min10cleanGL$step + 1), breaks = 100, main = "Step length en log", xlab = "log(step length + 1)",  xlim = c(0, 30))
  #1 mode around 0m nd one around 150m

hist(log(prep1min10cleanGL$step + 1e-3),   # avoid log(0)
     breaks = 100,
     main = "Step length (log-transformé) GL",
     xlab = "log(step length)",
     freq = FALSE)   # densité plutôt que fréquences
lines(density(log(prep1min10cleanGL$step + 1e-3)), col = "red", lwd = 2)
#Two modes: at 0 and 162m (exp(-1.8))

#look at density
x <- log(prep1min10cleanGL$step + 1)
d <- density(x, na.rm = TRUE)
plot(d,
     main = "Density of log(step length + 1)",
     xlab = "log(step length + 1)")

# local maximum
modes_idx <- which(diff(sign(diff(d$y))) == -2)

# modes on the log axis
modes_log <- d$x[modes_idx]
modes_log

modes_m <- exp(modes_log) - 1
modes_m #1.922605 186.927292

#-- TURNING ANGLES----
hist(prep1min10cleanGL$angle, breaks = seq(-pi, pi, length = 15), xlab = "angle", main = "")

hist(prep1min10cleanGL$angle,
     breaks = seq(-pi, pi, length = 30),
     xlab = "Turning angle (rad)",
     main = "Distribution des angles",
     freq = FALSE)

summary(prep1min10cleanGL$step)
 # Min.  1st Qu.   Median     Mean  3rd Qu.     Max. 
 #   0.000    1.846    4.654   73.432  124.383 1140.855 
summary(prep1min10cleanGL$angle)
#  Min.    1st Qu.     Median       Mean    3rd Qu.       Max.       NA's 
# -3.1415927 -1.9315975  0.0000232  0.0087748  1.9494781  3.1415927         16 

#B
#--STEP LENGTH----

# Plot histogram of step lengths 
hist(prep1min10cleanB$step, xlab = "step length (m)", main = "", breaks = 50, xlim = c(0, 500)) #in meters (UTM)
hist(log(prep1min10cleanB$step + 1), breaks = 100, main = "Step length en log", xlab = "log(step length + 1)",  xlim = c(0, 12))
#1 mode around 0m nd one around 150m

#look at density
hist(log(prep1min10cleanB$step + 1e-3), 
     breaks = 100,
     main = "Step length (log-transformed) B",
     xlab = "log(step length)",
     freq = FALSE)  
lines(density(log(prep1min10cleanB$step + 1e-3)), col = "red", lwd = 2)

#or
x <- log(prep1min10cleanB$step + 1)
d <- density(x, na.rm = TRUE)
plot(d,
     main = "Density of log(step length + 1)",
     xlab = "log(step length + 1)")

# local maximum
modes_idx <- which(diff(sign(diff(d$y))) == -2)

# modes on the log axis
modes_log <- d$x[modes_idx]
modes_log

modes_m <- exp(modes_log) - 1
modes_m #1.812388 95.474015

summary(prep1min10cleanB$angle)
#  Min.   1st Qu.    Median      Mean   3rd Qu.      Max.      NA's 
# -3.141593 -2.111989 -0.008801 -0.013795  2.083177  3.141593        23 

summary(prep1min10cleanB$step)
# Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#   0.000   1.756   4.340  69.776  70.112 994.516 

#-- TURNING ANGLES----
hist(prep1min10cleanB$angle, breaks = seq(-pi, pi, length = 15), xlab = "angle", main = "")

hist(prep1min10cleanB$angle,
     breaks = seq(-pi, pi, length = 30),
     xlab = "Turning angle (rad)",
     main = "Distribution des angles",
     freq = FALSE)

#Compare angles bteween study areas
df_angles <- bind_rows(
  prep1min10cleanGL %>% mutate(pop = "GL"),
  prep1min10cleanB  %>% mutate(pop = "B")
)

# Histograms
ggplot(df_angles, aes(x = angle, fill = pop)) +
  geom_histogram(
    aes(y = ..density..),
    breaks = seq(-pi, pi, length.out = 30),
    alpha = 0.5,
    position = "identity",
    color = "black"
  ) +
  scale_fill_manual(values = c("GL" = "steelblue", "B" = "orange")) +
  labs(
    x = "Turning angle (rad)",
    y = "Density",
    fill = "Study area
  "
  ) +
  theme_classic()

#density
ggplot(df_angles, aes(x = angle, fill = pop, colour = pop)) +
  geom_density(alpha = 0.4, linewidth = 0.6) +
  scale_x_continuous(
    limits = c(-pi, pi),
    breaks = c(-pi, -pi/2, 0, pi/2, pi),
    labels = c("-π", "-π/2", "0", "π/2", "π")
  ) +
  scale_y_continuous(limits = c(0, 0.35)) +
  scale_fill_manual(values = c("B" = "grey50", "GL" = "steelblue3")) +
  scale_colour_manual(values = c("B" = "grey30", "GL" = "steelblue4")) +
  labs(
    title = "Turning angle distributions",
    x = "Turning angle (rad)",
    y = "Density",
    fill = "Study area",
    colour = "Study area"
  ) +
  theme_minimal()


#Idem with step length
df_step <- bind_rows(
  prep1min10cleanGL %>% mutate(pop = "GL"),
  prep1min10cleanB  %>% mutate(pop = "B")
)

# Histograms
ggplot(df_step, aes(x = step, fill = pop)) +
  geom_histogram(
    aes(y = ..density..),
    alpha = 0.5,
    position = "identity",
    color = "black"
  ) +
  scale_fill_manual(values = c("GL" = "steelblue", "B" = "orange")) +
  labs(
    x = "Step length",
    y = "Density",
    fill = "Study area
  "
  ) +
  theme_classic()

ggplot(df_step, aes(x = step, fill = pop)) +
  geom_histogram(
    aes(y = ..density..),
    alpha = 0.5,
    position = "identity",
    color = "black",
  ) +
  scale_fill_manual(values = c("GL" = "steelblue", "B" = "orange")) +
  labs(
    x = "Step length",
    y = "Density",
    fill = "Study area
  "
  ) +
  theme_classic()

#density
ggplot(df_step, aes(x = step, fill = pop, colour = pop)) +
  geom_density(alpha = 0.4, linewidth = 0.6) +
  scale_x_continuous(
    limits = c(-0.3,15)
  ) +
  #scale_y_continuous(limits = c(0, 0.35)) +
  scale_fill_manual(values = c("B" = "grey50", "GL" = "steelblue3")) +
  scale_colour_manual(values = c("B" = "grey30", "GL" = "steelblue4")) +
  labs(
    title = "Step length (km), limited to 0-15km",
    x = "Step length (km)",
    y = "Density",
    fill = "Study area",
    colour = "Study area"
  ) +
  theme_minimal()

```

Models with 2-4-5 states not shown : follow the same process

## HMM GL 3 states, gamma vm -> not the optimal one
Initial values (fixed)

```{r}

stepMean0 <- c(m1 = 5, m2 =580, m3 = 167) # initial means (one for each state) 
stepSD0 <- c(5,580, 167 ) # initial standard deviations (one for each state) 
stepPar0 <- c(stepMean0, stepSD0)
#Distribution = usually Gamma

# Starting values for the turning angle parameters 
angleMean0 <- c(3.14, 0, 1.57) # initial means (one for each state) in radians (-pi;pi) 
angleCon0 <- c(rho1 = 0.1, rho2 = 0.8, rho = 0.2) # initial concentrations (one for each state) 
anglePar0 <- c(angleMean0, angleCon0)

#3 states model
stateNames <- c("resident","transit", "foraging")
zeromass <- c(0.0002, 0.0002, 0.0002) 

#with Von mises distribution for turning angles
Par0 <- list(step=c(stepMean0, stepSD0, zeromass), angle = c(angleMean0, angleCon0))
dist <- list(step = "gamma", angle = "vm")
#other possibility for turning angles "wrpcauchy"

fitted_modelGL <- fitHMM(data = prep1min10cleanGL, nbStates = 3, dist = dist, Par0 = Par0, stateNames = stateNames, estAngleMean = list(angle=TRUE), retryFits = 10)

print(fitted_modelGL) #likelihood (à maximiser) -210887

AIC(fitted_modelGL)

plot(fitted_modelGL, plotCI = TRUE)
#1 fitted distribution for the two states and total (gamma)
#2 wrapped distribution (concentration parameter)
#3 mapped states

#check if another state is needed
hmm_statesGL <- viterbi(fitted_modelGL)
#For a given model, reconstructs the most probable states sequence, using the Viterbi algorithm.

table(hmm_statesGL)/nrow(prep1min10cleanGL)
#globally

#from tuto mjones
# model fit can be assessed using the pseudo-residuals (e.g., residual autocorrelation fucntion plot below)
# compute pseudo-residuals for the steps and the angles
pr <- pseudoRes(fitted_modelGL)
# plot the ACF of step pseudo-residuals
acf(pr$stepRes[!is.na(pr$stepRes)],lag.max = 100) 
acf(pr$angleRes[is.finite(pr$angleRes)], lag.max = 100)#ok

#################### CHECK

prep1min10cleanGL$state <- hmm_statesGL
datastatesGL <- prep1min10cleanGL
datastatesGL$device <- sub("_.*", "", datastatesGL$ID) #extract device_id
datastatesGL$device_day <-  sub("^([^_]+_[^_]+).*", "\\1", datastatesGL$ID) #concatenate day and device_id

#proportion of each behaviour per individual_day, then averaged for the individual
  
# 1. 
etat_par_jour <- datastatesGL %>%
  group_by(device_day, device, state) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(device_day, device) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

# 2.
etat_moyen_par_individu <- etat_par_jour %>%
  group_by(device, state) %>%
  summarise(mean_prop = mean(prop), .groups = "drop") %>%
  arrange(device, state)

#Daily mean proportion of each state per individual
ggplot(etat_moyen_par_individu, aes(x = factor(device), y = mean_prop, fill = factor(state))) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    x = "Individual (device)",
    y = "Mean proportion",
    fill = "State"
  ) +
  theme_minimal()

# 3. Globally
etat_moyen_global <- etat_moyen_par_individu %>%
  group_by(state) %>%
  summarise(mean_prop_all_individuals = mean(mean_prop), .groups = "drop")

# add the study area indormation
datastatesall <- datastatesGL %>%
  mutate(pop = "GL")
#to later compare withe the brenne

################## Maps for some examples

# Select a sequence
id_select <- "243698_2024-06-11_5"

prep_id <- subset(datastatesGL, ID == id_select)

# Conversion sf to UTM 31N the, WGS84
prep_sf <- st_as_sf(prep_id, coords = c("x", "y"), crs = 32630) %>%
  st_transform(4326)

# Name the states
# stateNames <- c("Stationnaire", "Transit", "Foraging")
stateColors <- c("1" = "#F1C40F",
                 "3" = "#3498DB",
                 "2" = "#2ECC71")

# Add states
prep_sf$stateName <- factor(prep_sf$state,
                            levels = 1:3,
                            labels = c("1", "2", "3"))

# Line connecting points in chronological order
prep_line <- prep_sf %>%
  arrange(time) %>%
  summarise(do_union = FALSE) %>%
  st_cast("LINESTRING")

# Interactive mode
tmap_mode("view")

tm_shape(prep_line) +
  tm_lines(col = "grey60", lwd = 1.2) +
tm_shape(prep_sf) +
  tm_symbols(
    col = "stateName",
    palette = stateColors,
    size = 0.4,           
    border.lwd = 0.3,
    alpha = 0.9,
    popup.vars = c("État" = "stateName")
  ) +
  tm_layout(
    title = paste("Séquence", id_select),
    legend.outside = TRUE,
    frame = FALSE
  )

ggplot(prep1min10cleanGL, aes(x = step, fill = factor(hmm_statesGL))) +
  geom_density(alpha = 0.5) +
  scale_x_log10() +
  theme_minimal()

ggplot(prep1min10cleanGL, aes(x = angle, fill = factor(hmm_statesGL))) +
  geom_density(alpha = 0.5) +
  theme_minimal()


###### Try other sets of initial parameters
tryList <- list(
  list(step = c(5, 400, 120, 5, 400, 120, 0.0002, 0.0002, 0.0002),
       angle = c(3.14, 0, 1.57, 0.3, 0.8, 0.2)),
  list(step = c(10, 300, 150, 10, 300, 150, 0.0002, 0.0002, 0.0002),
       angle = c(3.14, 0, 1.57, 0.5, 0.8, 0.3))
)

fits <- lapply(tryList, function(Par0) {
  fitHMM(data = prep1min10cleanGL, nbStates = 3,
         dist = list(step = "gamma", angle = "vm"),
         Par0 = Par0, stateNames = stateNames,
         estAngleMean = list(angle = TRUE))
})


logLik_vec <- sapply(fits, function(fit) {
  if (!is.null(fit)) -fit$mod$minimum else NA
})

# AIC
AIC_vec <- sapply(fits, function(fit) {
  if (!is.null(fit)) AIC(fit) else NA
})

# Créer le dataframe
comparison_df <- data.frame(
  run = seq_along(fits),
  logLik = logLik_vec,
  AIC = AIC_vec
)

#Convergence to the same optimal parameters
```

### HMM GL 3 states, gamma WC --> Optimal one
Same initial values (fixed), change the distribution law for turning angles

```{r}

stepMean0 <- c(m1 = 5, m2 =580, m3 = 167) # initial means (one for each state) 
stepSD0 <- c(5,580, 167 ) # initial standard deviations (one for each state) 
stepPar0 <- c(stepMean0, stepSD0)
#Distribution = usually Gamma

# Starting values for the turning angle parameters 
angleMean0 <- c(3.14, 0, 1.57) # initial means (one for each state) in radians (-pi;pi) 
angleCon0 <- c(rho1 = 0.1, rho2 = 0.8, rho = 0.2) # initial concentrations (one for each state) 
anglePar0 <- c(angleMean0, angleCon0)

#3 states model
stateNames <- c("resident","transit", "foraging")
zeromass <- c(0.0002, 0.0002, 0.0002) 

#with Von mises distribution for turning angles
Par0 <- list(step=c(stepMean0, stepSD0, zeromass), angle = c(angleMean0, angleCon0))
dist <- list(step = "gamma", angle = "wrpcauchy")
#other possibility for turning angles "wrpcauchy"

fitted_modelGL2 <- fitHMM(data = prep1min10cleanGL, nbStates = 3, dist = dist, Par0 = Par0, stateNames = stateNames, estAngleMean = list(angle=TRUE), retryFits = 10)

print(fitted_modelGL2) #likelihood (maximise)

AIC(fitted_modelGL2)

plot(fitted_modelGL2, plotCI = TRUE)
#1 fitted distribution for the two states and total (gamma)
#2 wrapped distribution (concentration parameter)
#3 mapped states

#check if another state is needed
hmm_statesGL2 <- viterbi(fitted_modelGL2)
#For a given model, reconstructs the most probable states sequence, using the Viterbi algorithm.

table(hmm_statesGL2)/nrow(prep1min10cleanGL)

#from tuto mjones
# model fit can be assessed using the pseudo-residuals (e.g., residual autocorrelation fucntion plot below)
# compute pseudo-residuals for the steps and the angles
pr <- pseudoRes(fitted_modelGL2)
# plot the ACF of step pseudo-residuals
acf(pr$stepRes[!is.na(pr$stepRes)],lag.max = 100) 
acf(pr$angleRes[is.finite(pr$angleRes)], lag.max = 100)

#################### CHECK
prep1min10cleanGL2 <- prep1min10cleanGL
prep1min10cleanGL2$state <- hmm_statesGL2
datastatesGL2 <- prep1min10cleanGL2
datastatesGL2$device <- sub("_.*", "", datastatesGL2$ID) #extract device_id
datastatesGL2$device_day <-  sub("^([^_]+_[^_]+).*", "\\1", datastatesGL2$ID)

# 1. 
etat_par_jour2 <- datastatesGL2 %>%
  group_by(device_day, device, state) %>% 
  summarise(n = n(), .groups = "drop") %>%
  group_by(device_day, device) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

# 2. 
etat_moyen_par_individu2 <- etat_par_jour2GL %>%
  group_by(device, state) %>%
  summarise(mean_prop = mean(prop), .groups = "drop") %>%
  arrange(device, state)

ggplot(etat_moyen_par_individu2, aes(x = factor(device), y = mean_prop, fill = factor(state))) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    x = "Individu (device)",
    y = "Proportion moyenne",
    fill = "État comportemental"
  ) +
  theme_minimal()

# 3. globally
etat_moyen_global2 <- etat_moyen_par_individu2 %>%
  group_by(state) %>%
  summarise(mean_prop_all_individuals = mean(mean_prop), .groups = "drop")

# add study area information
datastatesall2 <- datastatesGL2 %>%
  mutate(pop = "GL")

#Maps for some examples
id_select <- "243698_2024-06-11_5"

prep_id <- subset(datastatesGL2, ID == id_select)

# Conversion sf to UTM 31N then WGS84
prep_sf <- st_as_sf(prep_id, coords = c("x", "y"), crs = 32630) %>%
  st_transform(4326)

stateNames <- c("Stationnaire", "Transit", "Foraging")
stateColors <- c("Stationnaire" = "#F1C40F",
                 "Transit" = "#3498DB",
                 "Foraging" = "#2ECC71")

prep_sf$stateName <- factor(prep_sf$state,
                            levels = 1:3,
                            labels = stateNames)

prep_line <- prep_sf %>%
  arrange(time) %>%
  summarise(do_union = FALSE) %>%
  st_cast("LINESTRING")


tmap_mode("view")
tm_shape(prep_line) +
  tm_lines(col = "grey60", lwd = 1.2) +
tm_shape(prep_sf) +
  tm_symbols(
    col = "stateName",
    palette = stateColors,
    size = 0.4,           # plus gros points visibles
    border.lwd = 0.3,
    alpha = 0.9,
    popup.vars = c("État" = "stateName")
  ) +
  tm_layout(
    title = paste("Séquence", id_select),
    legend.outside = TRUE,
    frame = FALSE
  )


### With previous and following sequences

id_select <- "243700_2024-06-12_10"
id_select <- "243698_2024-06-16_10"

id_root <- sub("_[0-9]+$", "", id_select)

prep_id <- datastatesGL2 %>%
  dplyr::filter(grepl(id_root, ID))

# Conversion sf to UTM 31N the, WGS84
prep_sf <- st_as_sf(prep_id, coords = c("x", "y"), crs = 32630) %>%
  st_transform(4326)

# Name the states
# stateNames <- c("Stationnaire", "Transit", "Foraging")
stateColors <- c("1" = "#F1C40F",
                 "3" = "#3498DB",
                 "2" = "#2ECC71")

# Add states
prep_sf$stateName <- factor(prep_sf$state,
                            levels = 1:3,
                            labels = c("1", "3", "2"))

# Line connecting points in chronological order
prep_line <- prep_sf %>%
  arrange(time) %>%
  summarise(do_union = FALSE) %>%
  st_cast("LINESTRING")

# Interactive mode
tmap_mode("view")

tm_shape(prep_line) +
  tm_lines(col = "grey60", lwd = 1.2) +
tm_shape(prep_sf) +
  tm_symbols(
    col = "stateName",
    palette = stateColors,
    size = 0.4,           
    border.lwd = 0.3,
    alpha = 0.9,
    popup.vars = c("État" = "stateName")
  ) +
  tm_layout(
    title = paste("Séquence", id_select),
    legend.outside = TRUE,
    frame = FALSE
  )


#density plots
ggplot(prep1min10cleanGL2, aes(x = step, fill = factor(hmm_statesGL2))) +
  geom_density(alpha = 0.5) +
  scale_x_log10() +
  theme_classic()

ggplot(prep1min10cleanGL2, aes(x = angle, fill = factor(hmm_statesGL2))) +
  geom_density(alpha = 0.5) +
  theme_classic()

#probably:
#red = stationnary
#blue = ARS
#green = extensive movement
resu <- datastatesGL2 %>% group_by(state) %>% summarize(mean_step = mean(step), sd_step = sd(step), mean_angle = mean(angle), sd_angle = sd(angle))

```

##### Additionnal representations
Plot on satellite map

```{r}
#plot eeach individual-day

# ID sélectionné
id_select <- "243697_2024-06-16_7"
id_root <- sub("_.*", "", id_select)
prep_id <- datastatesGL2 %>%
  dplyr::filter(grepl(id_root, ID))

prep_id$day <- sub(".*_(\\d{4}-\\d{2}-\\d{2})_.*", "\\1", prep_id$ID)

counts <- prep_id %>%
  count(day)

prep_id <- prep_id %>%
  left_join(counts, by = "day")

prep_id$day_label <- paste0(prep_id$day, " (n = ", prep_id$n, ")")

prep_id$day_label <- factor(
  prep_id$day_label,
  levels = unique(prep_id$day_label[order(prep_id$day)])
)

# Transformer en sf
prep_sf <- st_as_sf(prep_id, coords = c("x", "y"), crs = 32630) %>%
  st_transform(4326)

prep_sf$state <- as.character(prep_sf$state)
prep_sf$state <- as.factor(prep_sf$state)

# Mode interactif
tmap_mode("view")

# Carte
tm_shape(prep_sf) +
  tm_dots(
    col = "state",
    palette = c("1" = "#F1C40F",
                "2" = "#2ECC71",
                "3" = "#3498DB"),
    size = 0.4,
    alpha = 0.9
  ) +
  tm_facets(by = "day_label", free.coords = FALSE) +
  tm_layout(legend.outside = TRUE)

tm



###################### Final version Qgis

# --- 1. Select individual_day ---
id_select <- "243697_2024-06-10"

prep_id <- datastatesGL2 %>%
  filter(device_day == id_select) %>%
  arrange(time)

# --- 2. SEGMENTS (t → t+1) ---
segments <- prep_id %>%
  mutate(
    x_end = lead(x),
    y_end = lead(y),
    state_end = lead(state),
    time_end = lead(time)
  ) %>%
  filter(!is.na(x_end))

# --- 3. LINES sf ---
lines_sf <- segments %>%
  rowwise() %>%
  mutate(
    geometry = st_sfc(
      st_linestring(
        matrix(c(x, y, x_end, y_end), ncol = 2, byrow = TRUE)
      ),
      crs = 32630
    )
  ) %>%
  st_as_sf() %>%
  st_transform(4326)

# --- 4. POINTS sf ---
points_sf <- st_as_sf(prep_id, coords = c("x", "y"), crs = 32630) %>%
  st_transform(4326)

# --- 5. For QGIS ---

# Points : state en facteur/texte
points_sf$state_f <- as.factor(points_sf$state)

# Lignes : state de départ + arrivée
lines_sf$state_start <- as.factor(lines_sf$state)
lines_sf$state_end   <- as.factor(lines_sf$state_end)
lines_sf$segment_id <- 1:nrow(lines_sf)


tmap_mode("view")

tm_shape(lines_sf) +
  tm_lines(
    col = "state_end",   # état du point d’arrivée
    palette = stateColors,
    lwd = 1.2
  ) +
tm_shape(points_sf) +
  tm_symbols(
    fill = "state",
    fill.scale = tm_scale(values = stateColors),
    size = 0.4,
    border.lwd = 0.3,
    fill_alpha = 0.9
  ) +
tm_layout(
  title = id_select,
  legend.outside = TRUE,
  frame = FALSE
)

```

### HMM B 3 states, gamma vm ->  not the optimal one
Same initial values (fixed)

```{r}

stepMean0 <- c(m1 = 5, m2 =580, m3 = 167) # initial means (one for each state) 
stepSD0 <- c(5,580, 167 ) # initial standard deviations (one for each state) 
stepPar0 <- c(stepMean0, stepSD0)
#Distribution = usually Gamma

# Starting values for the turning angle parameters 
angleMean0 <- c(3.14, 0, 1.57) # initial means (one for each state) in radians (-pi;pi) 
angleCon0 <- c(rho1 = 0.1, rho2 = 0.8, rho = 0.2) # initial concentrations (one for each state) 
anglePar0 <- c(angleMean0, angleCon0)

#3 states model
stateNames <- c("resident","transit", "foraging")
zeromass <- c(0.0002, 0.0002, 0.0002) 

#with Von mises distribution for turning angles
Par0 <- list(step=c(stepMean0, stepSD0, zeromass), angle = c(angleMean0, angleCon0))
dist <- list(step = "gamma", angle = "vm")
#other possibility for turning angles "wrpcauchy"

fitted_modelB <- fitHMM(data = prep1min10cleanB, nbStates = 3, dist = dist, Par0 = Par0, stateNames = stateNames, estAngleMean = list(angle=TRUE), retryFits = 10)

print(fitted_modelB) #likelihood (à maximiser) -230741.5

AIC(fitted_modelB) #461528.9

plot(fitted_modelB, plotCI = TRUE)
#1 fitted distribution for the two states and total (gamma)
#2 wrapped distribution (concentration parameter)
#3 mapped states

#check if another state is needed
hmm_statesB <- viterbi(fitted_modelB)
#For a given model, reconstructs the most probable states sequence, using the Viterbi algorithm.

table(hmm_statesB)/nrow(prep1min10cleanB)

#from tuto mjones
# model fit can be assessed using the pseudo-residuals (e.g., residual autocorrelation fucntion plot below)
# compute pseudo-residuals for the steps and the angles
pr <- pseudoRes(fitted_modelB)
# plot the ACF of step pseudo-residuals
acf(pr$stepRes[!is.na(pr$stepRes)],lag.max = 100) #very nice
acf(pr$angleRes[is.finite(pr$angleRes)], lag.max = 100)#ok

#################### CHECK

prep1min10cleanB$state <- hmm_statesB
datastatesB <- prep1min10cleanB
datastatesB$device <- sub("_.*", "", datastatesB$ID) #extract device_id
datastatesB$device_day <-  sub("^([^_]+_[^_]+).*", "\\1", datastatesB$ID) #concatenate the day and device_id

#Compute the mean daily proportion of each state per individual
  
# 1. 
etat_par_joB <- datastatesB %>%
  group_by(device_day, device, state) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(device_day, device) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

# 2. 
etat_moyen_par_indiviB <- etat_par_joB %>%
  group_by(device, state) %>%
  summarise(mean_prop = mean(prop), .groups = "drop") %>%
  arrange(device, state)

ggplot(etat_moyen_par_indiviB, aes(x = factor(device), y = mean_prop, fill = factor(state))) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    x = "Individu (device)",
    y = "Proportion moyenne",
    fill = "État comportemental"
  ) +
  theme_minimal()

# 3. globally
etat_moyen_globB <- etat_moyen_par_indiviB %>%
  group_by(state) %>%
  summarise(mean_prop_all_individuals = mean(mean_prop), .groups = "drop")

datastatesB <- datastatesB %>% mutate(pop = "B")

# add the study area information
datastatesall <- bind_rows(datastatesall, datastatesB)


ggplot(prep1min10cleanB, aes(x = step, fill = factor(hmm_statesB))) +
  geom_density(alpha = 0.5) +
  scale_x_log10() +
  theme_minimal()

ggplot(prep1min10cleanB, aes(x = angle, fill = factor(hmm_statesB))) +
  geom_density(alpha = 0.5) +
  theme_minimal()

################## Maps for some examples

# Select a sequence
id_select <- "243695_2024-06-10_9"

prep_id <- subset(datastatesB, ID == id_select)

# Conversion sf to UTM 31N then WGS84
prep_sf <- st_as_sf(prep_id, coords = c("x", "y"), crs = 32631) %>%
  st_transform(4326)

# Name the states
stateNames <- c("Stationnaire", "Transit", "Foraging")
stateColors <- c("Stationnaire" = "#F1C40F",
                 "Transit" = "#3498DB",
                 "Foraging" = "#2ECC71")

# add the states
prep_sf$stateName <- factor(prep_sf$state,
                            levels = 1:3,
                            labels = stateNames)

prep_line <- prep_sf %>%
  arrange(time) %>%
  summarise(do_union = FALSE) %>%
  st_cast("LINESTRING")

tmap_mode("view")
tm_shape(prep_line) +
  tm_lines(col = "grey60", lwd = 1.2) +
tm_shape(prep_sf) +
  tm_symbols(
    col = "stateName",
    palette = stateColors,
    size = 0.4,           
    border.lwd = 0.3,
    alpha = 0.9,
    popup.vars = c("État" = "stateName")
  ) +
  tm_layout(
    title = paste("Séquence", id_select),
    legend.outside = TRUE,
    frame = FALSE
  )

```

### HMM B 3 states, gamma WC
Same initial values (fixed)

```{r}

stepMean0 <- c(m1 = 5, m2 =580, m3 = 167) # initial means (one for each state) 
stepSD0 <- c(5,580, 167 ) # initial standard deviations (one for each state) 
stepPar0 <- c(stepMean0, stepSD0)
#Distribution = usually Gamma

# Starting values for the turning angle parameters 
angleMean0 <- c(3.14, 0, 1.57) # initial means (one for each state) in radians (-pi;pi) 
angleCon0 <- c(rho1 = 0.1, rho2 = 0.8, rho = 0.2) # initial concentrations (one for each state) 
anglePar0 <- c(angleMean0, angleCon0)

#3 states model
stateNames <- c("resident","transit", "foraging")
zeromass <- c(0.0002, 0.0002, 0.0002) 

Par0 <- list(step=c(stepMean0, stepSD0, zeromass), angle = c(angleMean0, angleCon0))
dist <- list(step = "gamma", angle = "wrpcauchy")
#other possibility for turning angles "wrpcauchy"

fitted_modelB2 <- fitHMM(data = prep1min10cleanB, nbStates = 3, dist = dist, Par0 = Par0, stateNames = stateNames, estAngleMean = list(angle=TRUE), retryFits = 10)

print(fitted_modelB2)

AIC(fitted_modelB2) # 459829.8

plot(fitted_modelB2, plotCI = TRUE)

hmm_statesB2 <- viterbi(fitted_modelB2)
#For a given model, reconstructs the most probable states sequence, using the Viterbi algorithm.

table(hmm_statesB2)/nrow(prep1min10cleanB)

#from tuto mjones
# model fit can be assessed using the pseudo-residuals (e.g., residual autocorrelation fucntion plot below)
# compute pseudo-residuals for the steps and the angles
pr <- pseudoRes(fitted_modelB2)
# plot the ACF of step pseudo-residuals
acf(pr$stepRes[!is.na(pr$stepRes)],lag.max = 100) 
acf(pr$angleRes[is.finite(pr$angleRes)], lag.max = 100)

#################### CHECK
prep1min10cleanB2 <- prep1min10cleanB
prep1min10cleanB2$state <- hmm_statesB2
datastatesB2 <- prep1min10cleanB2
datastatesB2$device <- sub("_.*", "", datastatesB2$ID) #extraictdevice_id
datastatesB2$device_day <-  sub("^([^_]+_[^_]+).*", "\\1", datastatesB2$ID) #concatenate day and device_id

resu1 <- datastatesB2 %>% group_by(state) %>% summarize(mean_step = mean(step), sd_step = sd(step), mean_angle = mean(angle), sd_angle = sd(angle))
  
# 1. 
etat_par_jour2B <- datastatesB2 %>%
  group_by(device_day, device, state) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(device_day, device) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

# 2. 
etat_moyen_par_individu2 <- etat_par_jour2B %>%
  group_by(device, state) %>%
  summarise(mean_prop = mean(prop), .groups = "drop") %>%
  arrange(device, state)

ggplot(etat_moyen_par_individu2, aes(x = factor(device), y = mean_prop, fill = factor(state))) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    x = "Individual (device)",
    y = "Mean daily proportion",
    fill = "State"
  ) +
  theme_minimal()

# 3. globally
etat_moyen_global2 <- etat_moyen_par_individu2 %>%
  group_by(state) %>%
  summarise(mean_prop_all_individuals = mean(mean_prop), .groups = "drop")


datastatesB2 <- datastatesB2 %>% mutate(pop = "B")
datastatesall2 <- bind_rows(datastatesall2, datastatesB2)


################## Maps

#for some sequences: examples
id_select <- "243695_2024-06-10_9"

prep_id <- subset(datastatesB2, ID == id_select)

# Conversion sf to UTM 31N then WGS84
prep_sf <- st_as_sf(prep_id, coords = c("x", "y"), crs = 32631) %>%
  st_transform(4326)

# Labels (temporary)
stateNames <- c("Stationnaire", "Transit", "Foraging")
stateColors <- c("Stationnaire" = "#F1C40F",
                 "Transit" = "#3498DB",
                 "Foraging" = "#2ECC71")

prep_sf$stateName <- factor(prep_sf$state,
                            levels = 1:3,
                            labels = stateNames)

prep_line <- prep_sf %>%
  arrange(time) %>%
  summarise(do_union = FALSE) %>%
  st_cast("LINESTRING")

tmap_mode("view")

tm_shape(prep_line) +
  tm_lines(col = "grey60", lwd = 1.2) +
tm_shape(prep_sf) +
  tm_symbols(
    col = "stateName",
    palette = stateColors,
    size = 0.4,           
    border.lwd = 0.3,
    alpha = 0.9,
    popup.vars = c("État" = "stateName")
  ) +
  tm_layout(
    title = paste("Séquence", id_select),
    legend.outside = TRUE,
    frame = FALSE
  )

```


##### Additionnal representations

```{r}
stateColors <- c("Stationary" = "orange",
                 "Transit_flight" = "blue",
                 "ARS" = "darkgreen")

stateNames <- c("Stationary", "Transit_flight", "ARS")

# add labels
datastatesB2$stateName <- factor(datastatesB2$state,
                                       levels = 1:3,
                                       labels = stateNames)

# Conversion to sf with CRS UTM 31N (EPSG:32631)
prep_sf <- st_as_sf(datastatesB2,
                    coords = c("x", "y"),
                    crs = 32631)   # UTM zone 31N (Europe, WGS84)

# Transformation to WGS84 (lat/lon)
prep_sf <- st_transform(prep_sf, crs = 4326)

#add states
prep_sf$state <- datastatesB2$state
prep_sf$stateName <- factor(prep_sf$state,
                            levels = 1:3,
                            labels = stateNames)
#map
tmap_mode("view")  

tm_shape(prep_sf) +
  tm_symbols(col = "stateName",
             palette = stateColors,
             size = 0.2,
             alpha = 0.7,
             popup.vars = c("State" = "stateName"))

prep_line <- prep_sf %>%
  summarise(do_union = FALSE) %>%
  st_cast("LINESTRING")

tm_shape(prep_line) +
  tm_lines(col = "grey50") +
tm_shape(prep_sf) +
  tm_symbols(col = "stateName",
             palette = stateColors,
             size = 0.02,
             alpha = 0.7)

### With previous and following sequences

id_select <- "243695_2024-06-07_9"

id_root <- sub("_[0-9]+$", "", id_select)

prep_id <- datastatesB2 %>%
  dplyr::filter(grepl(id_root, ID))

# Conversion sf to UTM 31N the, WGS84
prep_sf <- st_as_sf(prep_id, coords = c("x", "y"), crs = 32631) %>%
  st_transform(4326)

# Name the states
# stateNames <- c("Stationnaire", "Transit", "Foraging")
stateColors <- c("1" = "#F1C40F",
                 "3" = "#3498DB",
                 "2" = "#2ECC71")

# Add states
prep_sf$stateName <- factor(prep_sf$state,
                            levels = 1:3,
                            labels = c("1", "2", "3"))

# Line connecting points in chronological order
prep_line <- prep_sf %>%
  arrange(time) %>%
  summarise(do_union = FALSE) %>%
  st_cast("LINESTRING")

# Interactive mode
tmap_mode("view")

tm_shape(prep_line) +
  tm_lines(col = "grey60", lwd = 1.2) +
tm_shape(prep_sf) +
  tm_symbols(
    col = "stateName",
    palette = stateColors,
    size = 0.4,           
    border.lwd = 0.3,
    alpha = 0.9,
    popup.vars = c("État" = "stateName")
  ) +
  tm_layout(
    title = paste("Séquence", id_select),
    legend.outside = TRUE,
    frame = FALSE
  )

###

id_select <- "243711_2024-06-12_7"
id_root <- sub("_[0-9]+$", "", id_select)

prep_id <- prep_sf %>%
  arrange(time) %>%
  mutate(
    segment_id = cumsum(state != lag(state, default = first(state)))
  )

# Conversion sf
prep_sf <- st_as_sf(prep_id, coords = c("x", "y"), crs = 32631) %>%
  st_transform(4326)

# add labels
datastatesB2$stateName <- factor(datastatesB2$state,
                                       levels = 1:3,
                                       labels = stateNames)

# États
stateColors <- c("1" = "#F1C40F",
                 "2" = "#2ECC71",
                 "3" = "#3498DB")

prep_sf$stateName <- factor(prep_sf$state,
                            levels = 1:3,
                            labels = c("1", "2", "3"))

# Ligne globale (optionnel)
prep_lines <- prep_id %>%
  arrange(time) %>%
  group_by(segment_id, stateName) %>%
  summarise(do_union = FALSE) %>%
  st_cast("LINESTRING")

# Visualisation
tmap_mode("view")

tm_shape(prep_lines) +
  tm_lines(
    col = "stateName",
    palette = stateColors,
    lwd = 1.2
  ) +
tm_shape(prep_id) +
  tm_symbols(
    fill = "stateName",
    fill.scale = tm_scale(values = stateColors),
    size = 0.4,
    border.lwd = 0.3,
    fill_alpha = 0.9
  ) +
tm_layout(
  legend.outside = TRUE,
  frame = FALSE
)

########
# --- 1. Sélection individu-jour ---
id_select <- "243708_2024-06-12"
prep_id <- datastatesB2 %>%
  filter(device_day == id_select) %>%
  arrange(time)

# --- 2. Créer les segments (points t → t+1) ---
segments <- prep_id %>%
  mutate(
    x_end = lead(x),
    y_end = lead(y),
    state_end = lead(state),
    time_end = lead(time)
  ) %>%
  filter(!is.na(x_end))  # enlever dernier point

# --- 3. Transformer en lignes sf ---
lines_sf <- segments %>%
  rowwise() %>%
  mutate(
    geometry = st_sfc(
      st_linestring(
        matrix(c(x, y, x_end, y_end), ncol = 2, byrow = TRUE)
      ),
      crs = 32631
    )
  ) %>%
  st_as_sf() %>%
  st_transform(4326)

# --- 4. Points sf ---
points_sf <- st_as_sf(prep_id, coords = c("x", "y"), crs = 32631) %>%
  st_transform(4326)

# --- 5. Couleurs ---
stateColors <- c("1" = "#F1C40F",
                 "2" = "#2ECC71",
                 "3" = "#3498DB")

# --- 6. Visualisation ---
tmap_mode("view")

tm_shape(lines_sf) +
  tm_lines(
    col = "state_end",   # état du point d’arrivée
    palette = stateColors,
    lwd = 1.2
  ) +
tm_shape(points_sf) +
  tm_symbols(
    fill = "state",
    fill.scale = tm_scale(values = stateColors),
    size = 0.4,
    border.lwd = 0.3,
    fill_alpha = 0.9
  ) +
tm_layout(
  title = id_select,
  legend.outside = TRUE,
  frame = FALSE
)



# --- 1. ---
id_select <- "243713_2024-06-16"

prep_id <- datastatesB2 %>%
  filter(device_day == id_select) %>%
  arrange(time)

# --- 2. SEGMENTS (t → t+1) ---
segments <- prep_id %>%
  mutate(
    x_end = lead(x),
    y_end = lead(y),
    state_end = lead(state),
    time_end = lead(time)
  ) %>%
  filter(!is.na(x_end))

# --- 3. LINES sf ---
lines_sf <- segments %>%
  rowwise() %>%
  mutate(
    geometry = st_sfc(
      st_linestring(
        matrix(c(x, y, x_end, y_end), ncol = 2, byrow = TRUE)
      ),
      crs = 32631
    )
  ) %>%
  st_as_sf() %>%
  st_transform(4326)

# --- 4. POINTS sf ---
points_sf <- st_as_sf(prep_id, coords = c("x", "y"), crs = 32631) %>%
  st_transform(4326)

# --- 5.For QGIS ---

points_sf$state_f <- as.factor(points_sf$state)

lines_sf$state_start <- as.factor(lines_sf$state)
lines_sf$state_end   <- as.factor(lines_sf$state_end)

lines_sf$segment_id <- 1:nrow(lines_sf)


tmap_mode("view")

tm_shape(lines_sf) +
  tm_lines(
    col = "state_end",   # état du point d’arrivée
    palette = stateColors,
    lwd = 1.2
  ) +
tm_shape(points_sf) +
  tm_symbols(
    fill = "state",
    fill.scale = tm_scale(values = stateColors),
    size = 0.4,
    border.lwd = 0.3,
    fill_alpha = 0.9
  ) +
tm_layout(
  title = id_select,
  legend.outside = TRUE,
  frame = FALSE
)

```

