---
title: "WTStats_tests_models"
author: "Oceane BEGASSAT"
date: "2026-05-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Load packages
```{r}
library(tidyverse)
library(gridExtra)
library(MuMIn) #pour AIcc
library(lme4) #pour glmer
library(emmeans)
library(performance)#R²
library(suncalc)  # getSunlightTimes()
```

## 1 ) Stats AKDEc

### Load + boxplots

```{r}
indiv_info2 <- readRDS( "dataMvmt_stage.rds")

#pivot to have a column "stage" as explicative variable
indiv_long_with <- indiv_info2 %>%
  dplyr::select(name, pop, colonie, AKDE95inc, AKDE95rea, , sex, pop) %>%
  pivot_longer(
    cols = c(AKDE95inc, AKDE95rea),
    names_to = "stage",
    values_to = "AKDE95"
  ) %>%
  mutate(
    stage = ifelse(stage == "AKDE95inc", "inc", "rea"),
  )

indiv_long_with$name <- as.factor(indiv_long_with$name)
indiv_long_with$stage <- as.factor(indiv_long_with$stage)
indiv_long_with$sex <- as.factor(indiv_long_with$sex)
indiv_long_with$colonie <- as.factor(indiv_long_with$colonie)

indiv_long <- indiv_long_with
```

### Effect of stage: Paired T test
from https://en.datainvestigatorth.com/post/the-difference-between-paired-sample-t-test-and-wilcoxon-signed-rank-test-when-to-use-and-how-they


Assumptions of the Paired Sample T-Test

The two sets of data come from the same group (dependent samples). yes
The differences between the paired values are normally distributed.
The dependent variable is measured on an interval or ratio scale. yes

The Wilcoxon Signed Rank Test is a non-parametric alternative to the Paired Sample T-Test.
It is used when the data are not normally distributed or contain outliers, making the mean unreliable as a measure of central tendency.

The choice depends mainly on the distribution and measurement level of your data.

If your data are normally distributed and quantitative, use the Paired Sample T-Test.
If your data are non-normal or contain outliers, use the Wilcoxon Signed Rank Test.
If your dependent variable is ordinal (e.g., ranked or rating data), use the Wilcoxon Signed Rank Test.
If your dependent variable is interval or ratio (e.g., scores, time, income), use the Paired Sample T-Test.

```{r}

#paired dataframe
paired <- indiv_long %>%
  dplyr::select(name, stage, AKDE95) %>%
  filter(stage %in% c("inc", "rea")) %>%
  tidyr::pivot_wider(
    names_from = stage,
    values_from = AKDE95,
    names_prefix = "AKDE_"
  ) %>%
  filter(!is.na(AKDE_inc) & !is.na(AKDE_rea))
#n = 10


#################### Global t.test

#is the difference between the two stages normally ditributed?
diff_ <- paired$AKDE_rea-paired$AKDE_inc
hist(diff_) #non normal
par(mfrow = c(1,2))
qqnorm(diff_); qqline(diff_) #no
shapiro.test(diff_) #W = 0.64691, p-value = 0.0002015

paired$AKDE_inclog <- log(paired$AKDE_inc)
paired$AKDE_realog <- log(paired$AKDE_rea)
diff_log <- paired$AKDE_realog-paired$AKDE_inclog
qqnorm(diff_log); qqline(diff_log)#ok 
hist(diff_log)
shapiro.test(diff_log) #W = 0.97808, p-value = 0.9541
#we ca use paired t test but only on log transformed data

paired$AKDE_incsqrt <- sqrt(paired$AKDE_inc )
paired$AKDE_reasqrt <- sqrt(paired$AKDE_rea)
diff_sq <- paired$AKDE_reasqrt-paired$AKDE_incsqrt
qqnorm(diff_sq); qqline(diff_sq)#ok mais pas très beau
hist(diff_sq) #pas beau
shapiro.test(diff_sq) #W = 0.84939, p-value = 0.05712 worst

qqnorm(diff_); qqline(diff_)
qqnorm(diff_log); qqline(diff_log) #better
qqnorm(diff_sq); qqline(diff_sq)

#On log scale
t.test(
  log(paired$AKDE_inc),
  log(paired$AKDE_rea),
  paired = TRUE
)

#or
t.test(diff_log)


############################ Separate by pop #######################################

paired <- indiv_long %>%
  dplyr::select(name, pop, stage, AKDE95) %>%
  filter(stage %in% c("inc", "rea")) %>%
  tidyr::pivot_wider(
    names_from = stage,
    values_from = AKDE95,
    names_prefix = "AKDE_"
  ) %>%
  filter(!is.na(AKDE_inc) & !is.na(AKDE_rea))

paired$AKDE_inclog <- log(paired$AKDE_inc )
paired$AKDE_realog <- log(paired$AKDE_rea)

gl <- paired %>% filter(pop == "GL")
b <- paired %>% filter(pop == "B")

#is the difference between the two stages normally ditributed?
diff_ <- gl$AKDE_rea-gl$AKDE_inc
hist(diff_) #non normal
par(mfrow = c(1,2))
qqnorm(diff_); qqline(diff_) #no
shapiro.test(diff_)#W = 0.9222, p-value = 0.5493

diff_ <- b$AKDE_rea-b$AKDE_inc
hist(diff_) #non normal
par(mfrow = c(1,2))
qqnorm(diff_); qqline(diff_) #no
shapiro.test(diff_)#W = 0.79642, p-value = 0.05455

diff_log <- gl$AKDE_realog-gl$AKDE_inclog
qqnorm(diff_log); qqline(diff_log)#ok 
hist(diff_log)
shapiro.test(diff_log)#W = 0.98898, p-value = 0.9522

diff_log <- b$AKDE_realog-b$AKDE_inclog
qqnorm(diff_log); qqline(diff_log)#ok 
hist(diff_log)
shapiro.test(diff_log) #W = 0.95307, p-value = 0.7651

t.test(
  log(gl$AKDE_inc),
  log(gl$AKDE_rea),
  paired = TRUE
)
#t = -1.6194, df = 3, p-value = 0.2038

t.test(
  log(b$AKDE_inc),
  log(b$AKDE_rea),
  paired = TRUE
)
#t = -1.4772, df = 5, p-value = 0.1997


#plot
paired %>% 
  pivot_longer(cols = c(AKDE_inc, AKDE_rea),
               names_to = "stage",
               values_to = "AKDE") %>%
  ggplot(aes(x = stage, y = AKDE, group = name, color = pop)) +
  geom_line(alpha = 0.5) +
  geom_point(size = 3) +
  facet_wrap(~ pop) +
  scale_x_discrete(labels = c("AKDE_inc" = "Incubation", "AKDE_rea" = "Rearing")) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(x = "Stage", y = "AKDE95", 
       title = "Within-individual change in AKDE between stages\nFaceted by population")

```

### Effect of the sex : Wilcoxon tests

```{r}

### ALL data pooled
## Can we use a t-test?
indivf <- indiv_long %>% filter(sex == "f")
indivm <- indiv_long %>% filter(sex == "m")
hist(indivf$AKDE95)
hist(indivm$AKDE95)
#not normal at all
# + non independant data : two lines per indivduals (incubation + chick rearing)

#divide into two stages
indivinc <- indiv_long %>% filter(stage == "inc")
indivrea <- indiv_long %>% filter(stage == "rea")
indivrea <- indivrea[!is.na(indivrea$AKDE95), ]
hist(indivinc$AKDE95)
hist(indivrea$AKDE95)
#not normal (right skewed)

#log-transformation ?
hist(log(indivinc$AKDE95)) #perhaps
hist(log(indivrea$AKDE95))
#Not normal  : non parametric test


#The Wilcoxon test is a non-parametric statistical test used to determine if there are significant differences between two related groups
# By default, the wilcox.test function in R omits pairs with missing values.

#Per stage
table(indivinc$sex, indivinc$name)
#M = 9, F = 6
table(indivrea$sex, indivrea$name)
#M = 5, F = 5

wilcox.test(AKDE95 ~ sex, data = indivinc) #W = 30, p-value = 0.7756 -> W = 28, p-value = 0.9546
wilcox.test(AKDE95 ~ sex, data = indivrea)#W = 8, p-value = 0.4206 -> W = 3, p-value = 0.05556

```

### Effect of the study area : Wilcoxon tests

```{r}
indivgl <- indiv_long %>% filter(pop == "GL")
indivb <- indiv_long %>% filter(pop == "B")
par(mfrow = c(2, 2))  
hist(indivb$AKDE95)
hist(indivgl$AKDE95)
hist(log(indivb$AKDE95)) #normal
hist(log(indivgl$AKDE95))
#not normal
#dependance structure

#Effect of pop
#test by stage
wilcox.test(AKDE95 ~ pop, data = indivinc)#W = 37, p-value = 0.2721 -> W = 39, p-value = 0.181
wilcox.test(AKDE95 ~ pop, data = indivrea)#W = 20, p-value = 0.1143 -> W = 16, p-value = 0.4762

#Boxplot
ggplot(indiv_long, aes(x = stage, y = AKDE95, color = sex)) +
  geom_boxplot(alpha = 0.3) +
  geom_jitter(width = 0.1, size = 1) +
  facet_wrap(~ pop) +
  theme_classic() +
  labs(
    x = "Stage",
    y = "AKDE95",
    title = "Home-range size"
  ) +
  theme(legend.position = "none")

###log transformed fo readibility

indiv_long$AKDE95log <- log(indiv_long$AKDE95)

ggplot(indiv_long, aes(x = stage, y = AKDE95log, fill = sex)) +
  geom_boxplot(
    outlier.shape = NA, show.legend = FALSE
  ) +
  geom_point(
    aes(colour = sex),
    position = position_jitterdodge(jitter.width = 0.20),
    show.legend = TRUE,
    size = 1,
    alpha = 0.7
  ) +
  facet_wrap(~ pop) +
  scale_fill_manual(values = c("white", "grey60")) +
  scale_colour_manual(values = c("white", "grey60")) +
  theme_classic()

ggplot(indiv_long, aes(x = stage, y = AKDE95log, colour = sex)) +
  geom_jitter(
    width = 0.1,
    size = 2,
    alpha = 0.9
  ) +
  facet_wrap(~ pop) +
  scale_colour_manual(values = c("grey50", "black")) +
  theme_classic()

#coloured (colorblind-proof)

g1 <- ggplot(indiv_long, aes(x = stage, y = AKDE95log, fill = sex)) +
  geom_boxplot(
    outlier.shape = NA, show.legend = TRUE
  ) +
  geom_point(
    aes(colour = sex),
    position = position_jitterdodge(jitter.width = 0.0),
    show.legend = F,
    size = 1,
    alpha = 0.7
  ) +
  facet_wrap(~ pop) +
  scale_fill_manual(values = c("#88ccee", "sienna2")) +
  scale_colour_manual(values = c("black", "black")) +
  theme_classic()

```

## 2 ) Stats daily distances

### Load
```{r}
stats_dist_day <- readRDS("data_Mvmt_day.rds")
#daily basis, n = 281

table(stats_dist_day$device_id, stats_dist_day$colonie)
stats_dist_day$colonie <- as.factor(stats_dist_day$colonie)

table(stats_dist_day$sex) 
table(stats_dist_day$pop)
table(stats_dist_day$stage) 
table(stats_dist_day$device_id) #3-39 days per individual
summary(stats_dist_day$n) #21-998 locs per day
```

### Y

Mean daily distance to the nest (only locs >12m from the nest)

```{r}
#Y distribution

par(mfrow=c(2,2))
# Boxplot
boxplot(stats_dist_day$mean_dist,col='blue',ylab='Daily mean distance from the nest')
# Cleveland plot
dotchart(stats_dist_day$mean_dist,pch=16,col='blue',xlab='Daily mean distance from the nest')
# Histogram
hist(stats_dist_day$mean_dist,col='blue',xlab="Daily mean distance from the nest",main="")
# Quantile-Quantile plot
qqnorm(stats_dist_day$mean_dist,pch=16,col='blue',xlab='')
qqline(stats_dist_day$mean_dist,col='red')
# Non normal
shapiro.test(stats_dist_day$mean_dist) #W = 0.76532, p-value = < 0.0001
#No outlier

```


### Xs

```{r}
par(mfrow=c(1,4))

# Sex
boxplot(mean_dist ~ sex,
        data = stats_dist_day,
        main = "Daily % of nesting bhv")
# pop
boxplot(
  mean_dist ~ pop,
  data = stats_dist_day,
  col = c('blue3', 'red2'),
  main = "Daily % of nesting bhv"
)

stage
boxplot(
  mean_dist ~ stage,
  data = stats_dist_day,
  col = c('blue3', 'red2'),
  main = "Daily % of nesting bhv"
)
summary(stats_dist_day$mean_dist)

```

### Choice of the random factors -> not transformed data

```{r}

#interactions ok ?

table(stats_dist_day$sex,stats_dist_day$stage )

table(stats_dist_day$sex,stats_dist_day$pop)

table(stats_dist_day$stage,stats_dist_day$pop)

#candidates : device, device and pop, pop, (colony )

#LM, constant model
mod0glm <- lm(mean_dist ~1,  data = stats_dist_day) #644.5083
#LMM constant avec random = indiv
mod0glmm <- lmer(mean_dist ~(1|device_id), data = stats_dist_day)
#LMM constant avec random = pop
mod0glmm1 <- lmer(mean_dist ~(1|pop),data = stats_dist_day) 
#LMM constant avec random = colonie
mod0glmm2 <- lmer(mean_dist ~(1|colonie), data = stats_dist_day) 
mod0glmm3 <- lmer(mean_dist ~(1|colonie) + (1|pop), data = stats_dist_day) 
#pop and colony do not carry variance


#LM complet
modglm <- lm(mean_dist ~ sex*stage ,  data = stats_dist_day)

#LMM complet avec random = indiv
modglmm <- lmer(mean_dist ~ sex*stage  + (1|device_id), data = stats_dist_day)

#LMM complet avec random = pop
modglmm1 <- lmer(mean_dist ~ sex*stage  + (1|pop), data = stats_dist_day) 

#LMM complet avec random = colony
modglmm3 <- lmer(mean_dist ~ sex*stage  + (1|colonie), data = stats_dist_day) 

#LMM complet avec random = pop+ device
modglmm2 <- lmer(mean_dist ~ sex*stage + (1|pop) +(1|device_id), data = stats_dist_day) 

#LMM complet avec random = device nested in pop
modglmm4 <- lmer(mean_dist ~ sex*stage + (1 | pop/device_id), data = stats_dist_day) 

#LMM complet avec random = device nested in colony
modglmm5 <- lmer(mean_dist ~ sex*stage + (1 | colonie/device_id), data = stats_dist_day) 

#Compare AICc
model_list <- list(
  mod0glm,
  mod0glmm,
  mod0glmm1,
  mod0glmm2,
  mod0glmm3
)
aiccs <- sapply(model_list, AICc)
aiccs#644.5083 556.8360 606.3602 563.1826 564.6317

model_list <- list(
  modglm,
  modglmm,
  modglmm1,
  modglmm3,
  modglmm2,
  modglmm4,
  modglmm5
)
aiccs <- sapply(model_list, AICc)
aiccs #640.9179 556.6524 613.6301 *551.9899* 557.4026 *550.3489*

best_model <- model_list[[ which.min(aiccs)]]
best_model #mean_dist ~ sex * stage + (1 | colonie/device_id)
#delta AICc <2 with only colonie more parsimonious
#But we will test with the nested structure (more accurate given the sampling method)

```


###  Model selection -> not transformed data
```{r}
#Selection with Dredge
fullmod <- lmer(
  mean_dist~ sex*stage + sex*pop + stage*pop + (1|colonie), data = stats_dist_day, na.action = na.fail, REML = F)
Dredge <- dredge(fullmod)
#all fixed factors retained, interaction pop*sex, sex*stage (unbalanced sample size)

modglm <- lmer(mean_dist~  sex*stage +
                  (1 | colonie), data = stats_dist_day, REML = F, na.action = na.fail)
AICc(modglm)
r.squaredGLMM(modglm)
confint(modglm, oldNames=FALSE)

#Standardized residuals vs predicted values: if random distribution = ok
plot(modglm, which = 1) #Heteroscedasticity ! Need a transformation
modglm1 <- lmer(mean_dist~  pop + sex + stage +  sex:stage + pop:sex +
                (1 | colonie), data = stats_dist_day, REML = F, na.action = na.fail)
AICc(modglm1)#541.9591
r.squaredGLMM(modglm)
plot(modglm1, which = 1) #Heteroskedasticity !
```

### Log transformation, model selection and validation

```{r}
stats_dist_day$logmeandist <- log(stats_dist_day$mean_dist)

#LM, constant model
mod0glm <- lm(logmeandist ~1,  data = stats_dist_day)
#LMM constant avec random = indiv
mod0glmm <- lmer(logmeandist ~(1|device_id), data = stats_dist_day)
#LMM constant avec random = pop
mod0glmm1 <- lmer(logmeandist ~(1|pop),data = stats_dist_day)
#LMM constant avec random = colonie
mod0glmm2 <- lmer(logmeandist ~(1|colonie), data = stats_dist_day)

mod0glmm3 <- lmer(logmeandist ~(1|colonie) + (1|device_id), data = stats_dist_day)
mod0glmm4 <- lmer(logmeandist ~(1|colonie) +(1|pop), data = stats_dist_day)
mod0glmm5 <- lmer(logmeandist ~(1|device_id) +(1|pop), data = stats_dist_day)
mod0glmm6 <- lmer(logmeandist ~(1|device_id) +(1|colonie), data = stats_dist_day)
mod0glmm7 <- lmer(logmeandist ~(1 | pop/device_id), data = stats_dist_day)
mod0glmm8 <- lmer(logmeandist ~(1 | colonie/device_id), data = stats_dist_day)


#LM complet
modglm <- lm(logmeandist ~ sex*stage + sex*pop + stage*pop ,  data = stats_dist_day)

#LMM complet avec random = indiv
modglmm <- lmer(logmeandist ~ sex*stage + sex*pop + stage*pop + (1|device_id), data = stats_dist_day)

#LMM complet avec random = pop
modglmm1 <- lmer(logmeandist ~ sex*stage + sex*pop + stage*pop + (1|pop), data = stats_dist_day)

#LMM complet avec random = colony
modglmm2 <- lmer(logmeandist ~ sex*stage + sex*pop + stage*pop + (1|colonie), data = stats_dist_day)

#LMM complet avec random = colony + device
modglmm3 <- lmer(logmeandist ~ sex*stage +sex*pop + stage*pop+ (1|pop) +(1|device_id), data = stats_dist_day)

#LMM complet avec random = device nested in colony
modglmm4 <- lmer(logmeandist ~ sex*stage + sex*pop + stage*pop+ (1 | pop/device_id), data = stats_dist_day)

modglmm5 <- lmer(logmeandist ~ sex*stage + sex*pop + stage*pop + (1|colonie) +(1|device_id), data = stats_dist_day)

modglmm6 <- lmer(logmeandist ~ sex*stage + sex*pop + stage*pop + (1 | colonie/device_id), data = stats_dist_day)

#Compare AICc
model_list <- list(mod0glm,
mod0glmm,
mod0glmm1,
mod0glmm2,
mod0glmm3,
mod0glmm4,
mod0glmm5,
mod0glmm6,
mod0glmm7 ,
mod0glmm8)
aiccs <- sapply(model_list, AICc)
aiccs #545.7701 404.9336 498.0159 444.5213 403.9124 445.7032 405.8515 403.9124 405.8515 403.9124

model_list <- list(
modglm,
modglmm,
modglmm1,
modglmm2,
modglmm3,
modglmm4,
modglmm5,
modglmm6)
aiccs <- sapply(model_list, AICc)
aiccs #491.0367 410.3688 514.6084 430.9587 412.5194 412.5194 407.2003 407.2003
best_model <- model_list[[ which.min(aiccs)]]
best_model #logmeandist ~ (1 | colonie) + (1 | device_id) 
#Equal with (1 | colonie/device_id) (more accurate for our sampling method)

##################

#Selection with Dredge
fullmod <- lmer(
  logmeandist~ sex*stage + sex*pop + stage*pop + (1 | colonie/device_id), data = stats_dist_day, na.action = na.fail, REML = F)
Dredge <- dredge(fullmod)
mods <- get.models(Dredge, subset = 1:10)


# data.frame
r2_df <- data.frame(
  Model = names(mods),
  do.call(
    rbind,
    lapply(mods, function(m) {
      as.data.frame(r2(m))
    })
  ),
  row.names = NULL
)

r2_df

#3 competitive models 
modglm1 <- lmer(logmeandist~  pop + sex + stage + sex:stage +  (1 | colonie/device_id), data = stats_dist_day, REML = F, na.action = na.fail)
AICc(modglm1) #392.2631
r.squaredGLMM(modglm1)
confint(modglm1)

#remove pop
modglm2 <- lmer(logmeandist~  sex + stage + sex:stage +  (1 | colonie/device_id), data = stats_dist_day, REML = F, na.action = na.fail)
AICc(modglm2) #392.444
r.squaredGLMM(modglm2)
confint(modglm2)

#add an interaction
modglm3 <- lmer(logmeandist~  pop + sex + stage + sex:stage + pop:sex +  (1 | colonie/device_id), data = stats_dist_day, REML = F, na.action = na.fail)
AICc(modglm3)#393.3435
r.squaredGLMM(modglm3)
confint(modglm3)

#adding the interaction do not improve the model a lot (NS)

# Top modèle : 2nd modèle
modglm <- lmer(logmeandist~ sex + stage + sex:stage +  (1 | colonie/device_id), data = stats_dist_day, REML = T, na.action = na.fail)
AICc(modglm)
r.squaredGLMM(modglm)
confint(modglm, oldNames=FALSE) 

#Standardized residuals vs predicted values: if random distribution = ok
plot(modglm, which = 1) #ok!

####### try with nested structure : qualitatively similar?
modglm <- lmer(logmeandist~  sex + stage + sex:stage + (1 | colonie/device_id), data = stats_dist_day, REML = T, na.action = na.fail)
AICc(modglm)
rsq(modglm)
r.squaredGLMM(modglm) #exactly the same
confint(modglm, oldNames=FALSE) #identical
modeglm <-  lmer(logmeandist~  sex + stage + sex:stage + (1 | colonie/device_id), data = stats_dist_day, REML = T, na.action = na.fail)
```

### Assumption checking

Residuals analysis

```{r}
mod1 <- modglm 
resid<-residuals(mod1, type="pearson")

par(mfrow=c(3,2))
# Histogram
hist(resid,col='blue',xlab="residuals",main="")
# Quantile-Quantile plot
qqnorm(resid,pch=16,col='blue',xlab='')
qqline(resid,col='red')

# residuals vs fitted
plot(resid~fitted(mod1)
      , col='blue'
      , pch=16)
abline(h = 0)
#ok!

# residuals against Sex
boxplot(resid~ stats_dist_day$sex,
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "Sex",
         main = "")
abline(h = 0)
#ok !


boxplot(resid~ stats_dist_day$stage, 
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "Stage",
         main = "")
abline(h = 0)
#ok

```

### Graphs and post-hoc

Post-hoc : see also help from https://aosmith.rbind.io/2019/03/25/getting-started-with-emmeans/
and https://cran.r-project.org/web/packages/emmeans/vignettes/AQuickStart.html

```{r}

g2 <- ggplot(stats_dist_day, aes(x = stage, y = logmeandist, fill = sex)) +
  geom_boxplot(
    outlier.shape = NA, show.legend = FALSE
  ) +
  geom_point(
    aes(colour = sex),
    position = position_jitterdodge(jitter.width = 0.20),
    show.legend = F,
    size = 0.5,
    alpha = 0.7
  ) +
  scale_fill_manual(values = c("#88ccee", "sienna2")) +
  scale_colour_manual(values = c("black", "black")) +
  theme_classic()


gx <- ggplot(stats_dist_day, aes(x = sex, y = logmeandist, fill = stage)) +
  geom_boxplot(
    outlier.shape = NA
  ) +
  geom_point(
    aes(colour = stage),
    position = position_jitterdodge(jitter.width = 0.20),
    show.legend = TRUE,
    size = 0.6,
    alpha = 0.7
  ) +
  scale_fill_manual(values = c("#88ccee", "sienna2")) +
  scale_colour_manual(values = c("black", "black")) +
  theme_classic()

emm <- emmeans(mod1, ~ sex * stage)
emm_df <- as.data.frame(emm)
ggplot(stats_dist_day, aes(x = interaction(stage, sex),
                          y = logmeandist)) +
  geom_boxplot(outlier.shape = NA, fill = "grey80") +
  geom_point(position = position_jitter(width = 0.1),
             size = 0.5, alpha = 0.5) +
  geom_point(data = emm_df,
             aes(x = interaction(stage, sex), y = emmean),
             colour = "red", size = 3) +
  theme_classic()

#Post-hoc: interaction

# Compute estimated marginal means (EMMs) for specified factors or factor combinations in a linear model; and optionally, comparisons or contrasts among them. EMMs are also known as least-squares means.

#see help here : https://bookdown.org/dereksonderegger/571/4-contrasts.html
#voir aussi https://www.statforbiology.com/2023/stat_emmeans_backtransform/#:~:text=It%20is%20very%20simple%3A%20emmeans%20auto-detects%20the%20transformation,that%20I%20will%20discuss%20in%20a%20future%20post%29.

mod1 <- lmer(logmeandist~ sex + stage + sex:stage +  (1 | colonie/device_id), data = stats_dist_day, REML = T, na.action = na.fail)
emm <- emmeans(mod1, ~ sex * stage, type="response")
emm

modglm <- lmer(log(mean_dist)~ sex + stage + sex:stage +  (1 | colonie/device_id), data = stats_dist_day, REML = T, na.action = na.fail)
emm <- emmeans(modglm, ~ sex * stage, type="response")
# préférable

contrast(emm, method = "pairwise", by = "stage", type = "response")
contrast(emm, method = "pairwise", by = "sex", type = "response")
contrast(emm, method = "revpairwise", by = "sex", type = "response")
 
dt <- as.data.frame(emm)
dt$emmean_km <- exp(dt$emmean)
dt$low <- exp(dt$lower.CL)
dt$upp <- exp(dt$upper.CL)

######

par(mfrow =c(1,2))
emmip(modglm, stage ~ sex, CIs = T)
emmip(modglm, sex ~ stage, CIs = T)

# 
pairs(emm)
contrast(emm, method = "revpairwise", by = "sex") #intra sex compare stage (rear vs incub et non  l'inverse)
contrast(emm, method = "revpairwise", by = "pop")

##### Quantify the difference

#Is there a differece between sexes intra-stage ?

modglm <- lmer(log(mean_dist)~ sex + stage + sex:stage +  (1 | colonie/device_id), data = stats_dist_day, REML = T, na.action = na.fail)
emmeans(modglm, ~ sex | stage, type = "response") %>% pairs()

#is there a difference between stages intra-sex ?
modglm <- lmer(log(mean_dist)~ sex + stage + sex:stage +  (1 | colonie/device_id), data = stats_dist_day, REML = T, na.action = na.fail)
emmeans(modglm, ~ stage | sex, type = "response") %>% pairs()

```

##3 ) Daily movement intensity
Daily cumulatve distances / delta t (km/h)

### Load 

```{r}
stats_dist_day <- readRDS ("data_Mvmt_day.rds")

table(stats_dist_day$sex) 
table(stats_dist_day$pop) 
table(stats_dist_day$stage) 
table(stats_dist_day$device_id) #3-39 days per individual
summary(stats_dist_day$n) #21-998

par(mfrow=c(1,3))

# Sex
boxplot(vitesse_kmh ~ sex,
        data = stats_dist_day,
        main = "")
# pop
boxplot(
 vitesse_kmh ~ pop,
  data = stats_dist_day,
  col = c('blue3', 'red2'),
  main = ""
)
#no difference

#stage
boxplot(
  vitesse_kmh ~ stage,
  data = stats_dist_day,
  col = c('blue3', 'red2'),
  main = ""
)
summary(stats_dist_day$vitesse_kmh)
```

### Y

```{r}
#Y distribution

par(mfrow=c(2,2))
# Boxplot
boxplot(stats_dist_day$vitesse_kmh,col='blue',ylab='Daily mean speed')
# Cleveland plot
dotchart(stats_dist_day$vitesse_kmh,pch=16,col='blue',xlab='Daily mean speed')
# Histogram
hist(stats_dist_day$vitesse_kmh,col='blue',xlab="Daily mean speed",main="")
# Quantile-Quantile plot
qqnorm(stats_dist_day$vitesse_kmh,pch=16,col='blue',xlab='')
qqline(stats_dist_day$vitesse_kmh,col='red')
# Non normal
shapiro.test(stats_dist_day$vitesse_kmh) #W = 0.92282, p-value = 6.999e-11
#No outlier

```

NOT shown : model with untransformed data -> heteroskedasticity

### Square root transformation

```{r}

stats_dist_day$sqvitesse_kmh <- sqrt(stats_dist_day$vitesse_kmh)

#Y distribution

par(mfrow=c(2,2))
# Boxplot
boxplot(stats_dist_day$sqvitesse_kmh,col='blue',ylab='Daily mean speed')
# Cleveland plot
dotchart(stats_dist_day$sqvitesse_kmh,pch=16,col='blue',xlab='Daily mean speed')
# Histogram
hist(stats_dist_day$sqvitesse_kmh,col='blue',xlab="Daily mean speed",main="")
# Quantile-Quantile plot
qqnorm(stats_dist_day$sqvitesse_kmh,pch=16,col='blue',xlab='')
qqline(stats_dist_day$sqvitesse_kmh,col='red')
# Non normal
shapiro.test(stats_dist_day$sqvitesse_kmh) 
#No outlier

```

### Xs
```{r}
# Sex
boxplot(sqvitesse_kmh ~ sex,
        data = stats_dist_day,
        main = "")
# pop
boxplot(
 sqvitesse_kmh ~ pop,
  data = stats_dist_day,
  col = c('blue3', 'red2'),
  main = ""
)
#no difference

#stage
boxplot(
  sqvitesse_kmh ~ stage,
  data = stats_dist_day,
  col = c('blue3', 'red2'),
  main = ""
)
```

### Model selection

```{r}


#LM, constant model
mod0glm <- lm(sqvitesse_kmh ~1,  data = stats_dist_day)
#LMM constant avec random = indiv
mod0glmm <- lmer(sqvitesse_kmh ~(1|device_id), data = stats_dist_day)
#LMM constant avec random = pop
mod0glmm1 <- lmer(sqvitesse_kmh ~(1|pop),data = stats_dist_day)
#LMM constant avec random = colonie
mod0glmm2 <- lmer(sqvitesse_kmh ~(1|colonie), data = stats_dist_day)

mod0glmm3 <- lmer(sqvitesse_kmh ~(1|colonie) + (1|device_id), data = stats_dist_day)
mod0glmm4 <- lmer(sqvitesse_kmh ~(1|colonie) +(1|pop), data = stats_dist_day)
mod0glmm5 <- lmer(sqvitesse_kmh ~(1|device_id) +(1|pop), data = stats_dist_day)
mod0glmm6 <- lmer(sqvitesse_kmh ~(1|device_id) +(1|colonie), data = stats_dist_day)
mod0glmm7 <- lmer(sqvitesse_kmh ~(1 | pop/device_id), data = stats_dist_day)
mod0glmm8 <- lmer(sqvitesse_kmh ~(1 | colonie/device_id), data = stats_dist_day)


#LM complet
modglm <- lm(sqvitesse_kmh ~ sex*stage + sex*pop + stage*pop ,  data = stats_dist_day)

#LMM complet avec random = indiv
modglmm <- lmer(sqvitesse_kmh ~ sex*stage + sex*pop + stage*pop + (1|device_id), data = stats_dist_day)

#LMM complet avec random = pop
# modglmm1 <- lmer(sqvitesse_kmh ~ sex*stage + sex*pop + stage*pop + (1|pop), data = stats_dist_day)

#LMM complet avec random = colony
modglmm2 <- lmer(sqvitesse_kmh ~ sex*stage + sex*pop + stage*pop + (1|colonie), data = stats_dist_day)

#LMM complet avec random = colony + device
# modglmm3 <- lmer(sqvitesse_kmh ~ sex*stage +sex*pop + stage*pop+ (1|pop) +(1|device_id), data = stats_dist_day)

#LMM complet avec random = device nested in colony
# modglmm4 <- lmer(sqvitesse_kmh ~ sex*stage + sex*pop + stage*pop+ (1 | pop/device_id), data = stats_dist_day)

modglmm5 <- lmer(sqvitesse_kmh ~ sex*stage + sex*pop + stage*pop + (1|colonie) +(1|device_id), data = stats_dist_day)

modglmm6 <- lmer(sqvitesse_kmh ~ sex*stage + sex*pop + stage*pop + (1 | colonie/device_id), data = stats_dist_day)

#Compare AICc
model_list <- list(
mod0glm,
mod0glmm,
mod0glmm1,
mod0glmm2,
mod0glmm3,
mod0glmm4,
mod0glmm5,
mod0glmm6,
mod0glmm7 ,
mod0glmm8)
aiccs <- sapply(model_list, AICc)
aiccs #348.5433 276.1129 355.9065 289.1370 274.6231 291.1953 277.8522 274.6231 277.8522 274.6231
  
  
model_list <- list(
modglm,
modglmm,
# modglmm1,
modglmm2,
# modglmm3,
# modglmm4,
modglmm5,
modglmm6)
aiccs <- sapply(model_list, AICc)
aiccs #296.4691 215.4444 235.8044 213.9351 213.9351
best_model <- model_list[[ which.min(aiccs)]]
best_model #sqvitesse_kmh ~ sex * stage + sex * pop + stage * pop + (1 |      colonie) + (1 | device_id)
#Equal with (1 | colonie/device_id) (more accurate for our sampling method)

##################

#Selection with Dredge
fullmod <- lmer(
  sqvitesse_kmh~ sex*stage + sex*pop + stage*pop + (1 | colonie/device_id), data = stats_dist_day, na.action = na.fail, REML = F)
Dredge <- dredge(fullmod)

modglm2 <- lmer(sqvitesse_kmh~  pop + stage + pop:stage +
                  (1 | colonie/device_id), data = stats_dist_day, REML = F, na.action = na.fail)
AICc(modglm2)#194.2543
r.squaredGLMM(modglm2)
confint(modglm2)

modglm2 <- lmer(sqvitesse_kmh~  pop + stage +sex +  pop:stage +
                  (1 | colonie/device_id), data = stats_dist_day, REML = F, na.action = na.fail)
AICc(modglm2)#195.0669
r.squaredGLMM(modglm2)
confint(modglm2)

modglm2 <- lmer(sqvitesse_kmh~  pop + stage+ sex + pop:stage + sex:stage +
                  (1 | colonie/device_id), data = stats_dist_day, REML = F, na.action = na.fail)
AICc(modglm2)#195.6341
r.squaredGLMM(modglm2)
confint(modglm2)

modglm2 <- lmer(sqvitesse_kmh~  pop + stage+ sex + pop:sex + pop:stage +
                  (1 | colonie/device_id), data = stats_dist_day, REML = F, na.action = na.fail)
AICc(modglm2)#196.3918
r.squaredGLMM(modglm2)
confint(modglm2)

mods <- get.models(Dredge, subset = 1:10)

library(performance)

# data.frame
r2_df <- data.frame(
  Model = names(mods),
  do.call(
    rbind,
    lapply(mods, function(m) {
      as.data.frame(r2(m))
    })
  ),
  row.names = NULL
)

r2_df


# Top model : stage:pop and stage
modglm <- lmer(sqvitesse_kmh~  stage + pop + stage:pop +
                  (1 | colonie/device_id), data = stats_dist_day, REML = T, na.action = na.fail)
AICc(modglm)#205.8624
r.squaredGLMM(modglm)
confint(modglm) 

#Standardized residuals vs predicted values: if random distribution = ok
plot(modglm, which = 1) #ok!
summary(modglm)
```

### Model validation: assumption checking

Residuals analysis

```{r}
mod1 <- modglm
resid<-residuals(mod1, type="pearson")

par(mfrow=c(3,2))
# Histogram
hist(resid,col='blue',xlab="residuals",main="")
# Quantile-Quantile plot
qqnorm(resid,pch=16,col='blue',xlab='')
qqline(resid,col='red')

# residuals vs fitted
plot(resid~fitted(mod1)
      , col='blue'
      , pch=16)
abline(h = 0)
#ok!


boxplot(resid~ stats_dist_day$stage, 
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "Stage",
         main = "")
abline(h = 0)
#ok


boxplot(resid~ stats_dist_day$pop, 
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "pop",
         main = "")
abline(h = 0)
#ok

```


### Post-hoc
```{r}

# Compute estimated marginal means (EMMs) for specified factors or factor combinations in a linear model; and optionally, comparisons or contrasts among them. EMMs are also known as least-squares means.

#see help here : https://bookdown.org/dereksonderegger/571/4-contrasts.html
# variation en % :² -> pas interprétable directement
#attention (mean(√y))^2 ≠ mean(y)

modglm <- lmer(sqrt(vitesse_kmh) ~  stage + pop + stage:pop +(1 | colonie/device_id), data = stats_dist_day, REML = T, na.action = na.fail)
emm <- emmeans(modglm, ~ stage*pop, type="response")
emm

emm2 = emmeans(modglm, specs = pairwise ~ pop|stage, type = "response")
emm2

contrast(emm, method = "revpairwise", by = "pop", type = "response")

emm_bt <- regrid(emm)   # maintenant contrasts sont en km/h
contrast(emm_bt, method="revpairwise", by="pop", type="response")
contrast(emm_bt, method="revpairwise", by="stage", type="response")

pairs(emm_bt)

# Compare stages within pop
emm <- emmeans(modglm, ~ stage| pop, type="response")
emm_bt <- regrid(emm)   
pairs(emm_bt)

emm <- emmeans(modglm, ~  pop|stage, type="response")
emm_bt <- regrid(emm)   
pairs(emm_bt)

emm_stage <- emmeans(
  modglm,         
  ~ stage,
  type = "response"
)
emm_stage

emm_stage_kmh <- summary(emm_stage)
emm_stage_kmh$estimate_kmh <- emm_stage_kmh$emmean^2
emm_stage_kmh$low_kmh <- emm_stage_kmh$lower.CL^2
emm_stage_kmh$upp_kmh <- emm_stage_kmh$upper.CL^2
emm_stage_kmh

diff_kmh <- emm_stage_kmh$estimate_kmh[emm_stage_kmh$stage == "rear"] -
            emm_stage_kmh$estimate_kmh[emm_stage_kmh$stage == "incub"]
diff_kmh 

perc_change <- (emm_stage_kmh$estimate_kmh[emm_stage_kmh$stage == "rear"] /
                emm_stage_kmh$estimate_kmh[emm_stage_kmh$stage == "incub"] - 1) * 100
perc_change 



```



### Représentation
```{r}

g3 <- ggplot(stats_dist_day, aes(x = stage, y = vitesse_kmh, fill = pop)) +
  geom_boxplot(
    outlier.shape = NA, show.legend = F
  ) +
  geom_point(
    aes(colour = pop),
    position = position_jitterdodge(jitter.width = 0.20),
    show.legend = F,
    size = 0.5,
    alpha = 0.7
  ) +
  scale_fill_manual(values = c("#88ccee", "sienna2")) +
  scale_colour_manual(values = c("black", "black")) +
  theme_classic()

gx <- ggplot(stats_dist_day, aes(x = pop, y = vitesse_kmh, fill = stage)) +
  geom_boxplot(
    outlier.shape = NA
  ) +
  geom_point(
    aes(colour = stage),
    position = position_jitterdodge(jitter.width = 0.20),
    show.legend = TRUE,
    size = 0.6,
    alpha = 0.7
  ) +
  scale_fill_manual(values = c("#88ccee", "sienna2")) +
  scale_colour_manual(values = c("black", "black")) +
  theme_classic()
```

##4 ) Daily time-activity budget
% of location in each of the three main behaviours identified

### Load data

```{r}
summary_states <-  readRDS("summary_statesdaily")
#275 x 18, summarizes the daily proportion of each of the three identified states


#In Incubation :
summary_days <- summary_states %>%
  filter(stage == "incub") %>%
  group_by(device) %>%
  summarise(
    nb_locs = n(),
    nb_days = n_distinct(time),
    deb = min(time),
    end = max(time),
    .groups = "drop"
  )

#Chick rearing
summary_days2 <- summary_states %>%
  filter(stage == "rear") %>%
  group_by(device) %>%
  summarise(
    nb_locs = n(),
    nb_days = n_distinct(time),
    deb = min(time),
    end = max(time),
    .groups = "drop"
  )

```

### Day coverage
To what extend do our locations cover the day?

```{r}

compute_day_length_summary <- function(df) {
  
  # 1️⃣Extract the dates
  date_loc_df <- df %>%
    mutate(date_only = as.Date(time)) %>%
    distinct(date_only, LAT_nid_R, LON_nid_R)
  
  # 2️⃣Sunset and sunlight
  sun_times_df <- date_loc_df %>%
    rowwise() %>%
    mutate(
      sun_times = list(getSunlightTimes(
        date = date_only,
        lat = LAT_nid_R,
        lon = LON_nid_R,
        tz = "UTC",
        keep = c("dawn", "dusk")
      ))
    ) %>%
    unnest_wider(sun_times) %>%
    mutate(
      day_length_min = as.numeric(difftime(dusk, dawn, units = "mins")),
      n_loc_per_min = round(day_length_min)  # 1 location / minute 
    ) %>%
    dplyr::select(date_only, LAT_nid_R, LON_nid_R, day_length_min, n_loc_per_min)
  
  # 3️⃣Merge
  df_out <- df %>%
    mutate(date_only = as.Date(time)) %>%
    left_join(sun_times_df, by = c("date_only", "LAT_nid_R", "LON_nid_R")) %>%
    dplyr::select(-date_only)
  
  return(df_out)
}

summary_states <- compute_day_length_summary(summary_states)
summary_states$prop_locsday <- summary_states$N_global/summary_states$n_loc_per_min*100
summary(summary_states$prop_locsday)

#Not shown: save and test different thresold of day coverage
#retained = 10%
summary_states_keep <- summary_states
summary_states <- summary_states %>% filter(prop_locsday>10)

summary_days5 <- summary_states %>%
  filter(stage == "incub") %>%
  group_by(device) %>%
  summarise(
    nb_locs = n(),
    .groups = "drop"
  )

summary_days6 <- summary_states %>%
  filter(stage == "rear") %>%
  group_by(device) %>%
  summarise(
    nb_locs = n(),
    .groups = "drop"
  )


valid_days <- summary_states %>%
  distinct(device_day)

summary_states <- summary_states %>% select(-c(endrepro, day_length_min, n_loc_per_min, prop_locsday))


table(summary_states$sex) 
table(summary_states$pop.x) 
table(summary_states$stage) 
table(summary_states$device) 
summary(summary_states$N_global) #103-968
```

 Not shown: remove the stationary locations <12m from the nest (= roosting)
 
### Split by study area
```{r}
summary_states2 <-  readRDS("summary_statesdailyclean")
summary_states2$device <- as.factor(summary_states2$device)
summary_states2$stage <- as.factor(summary_states2$stage)
summary_states2$nb_eggs <- as.factor(summary_states2$nb_eggs)


summary_statesGL <- summary_states2 %>% 
  filter(pop.x == "GL") %>% 
  droplevels()

summary_statesB <- summary_states2 %>% 
  filter(pop.x == "B") %>% 
  droplevels()

```

## Brenne

### %Stationary

```{r}
#Y distribution
#Y between 0 and 1 (%)
ratio5 <- cbind(summary_statesB$n_nesting , summary_statesB$N_global-summary_statesB$n_nesting ) #for the glm building  

par(mfrow=c(2,2))
# Boxplot
boxplot(summary_statesB$prop_nesting,col='blue',ylab='Daily %of resting at the nest')
# Cleveland plot
dotchart(summary_statesB$prop_nesting,pch=16,col='blue',xlab='Monthly mean \n of the daily %of resting at the nest')
# Histogram
hist(summary_statesB$prop_nesting,col='blue',xlab="Monthly mean\n of the daily %of resting at the nest",main="")
# Quantile-Quantile plot
qqnorm(summary_statesB$prop_nesting,pch=16,col='blue',xlab='')
qqline(summary_statesB$prop_nesting,col='red')
# Non normal
shapiro.test(summary_statesB$prop_nesting) #W = 0.94015, p-value = 0.0001067
#No outlier

```
### Xs
```{r}
par(mfrow=c(1,2))

# Sex
boxplot(prop_nesting ~ sex,
        data = summary_statesB,
        main = "Daily % of nesting bhv")

#stage
boxplot(
  prop_nesting ~ stage,
  data = summary_statesB,
  col = c('blue3', 'red2'),
  main = "Daily % of nesting bhv"
)


summary(summary_statesB$prop_nesting)

```

### Choice of the random factors

```{r}
#Interactions
table(summary_statesB$sex,summary_statesB$stage )

#GLM, constant model
mod0glm <- glm(ratio5~1, family = binomial(link = "logit"), data = summary_statesB)
#GLMM constant avec random = indiv
mod0glmm <- glmer(ratio5~(1|device),family = binomial(link = "logit"), data = summary_statesB)
#GLMM constant avec random = colony
mod0glmm1 <- glmer(ratio5~(1|colony),family = binomial(link = "logit"), data = summary_statesB)
#GLMM constant avec random = colony
mod0glmm2 <- glmer(ratio5~(1|colony),family = binomial(link = "logit"), data = summary_statesB)

#GLM complet
modglm <- glm(ratio5~ sex*stage, family = binomial(link = "logit"), data = summary_statesB)

#GLMM complet avec random = indiv
modglmm <- glmer(ratio5~ sex*stage + (1|device),family = binomial(link = "logit"), data = summary_statesB)

#GLMM complet avec random = colony
modglmm1 <- glmer(ratio5~ sex*stage + (1|colony),family = binomial(link = "logit"), data = summary_statesB)

#GLMM complet avec random = colony + device
modglmm2 <- glmer(ratio5~ sex*stage + (1|colony) +(1|device),family = binomial(link = "logit"), data = summary_statesB) #singular

#GLMM complet avec random = device nested in colony
modglmm3 <- glmer(ratio5~ sex*stage + (1 | colony/device), family = binomial(link = "logit"), data = summary_statesB) #singular

#Compare AICc
model_list <- list(mod0glm,mod0glmm, mod0glmm1, mod0glmm2, modglm, modglmm, modglmm1, modglmm2, modglmm3)
aiccs <- sapply(model_list, AICc)
aiccs 

model_list <- list(modglm, modglmm, modglmm1, modglmm2, modglmm3)
aiccs <- sapply(model_list, AICc)
aiccs
best_model <- model_list[[ which.min(aiccs)]]
best_model 

```

###  Model selection
```{r}
#Selection with Dredge
fullmod <- glmer(
  ratio5~ sex*stage + (1|device),family = binomial(link = "logit"), data = summary_statesB , na.action = na.fail)
Dredge <- dredge(fullmod)

mods <- get.models(Dredge, subset = 1:5)  
r2_list <- lapply(mods, r.squaredGLMM)
r2_df <- do.call(rbind, r2_list)
r2_df <- cbind(Model = names(mods), r2_df)
print(r2_df)

mod1 <- glmer(
  ratio5~ stage + (1|device),family = binomial(link = "logit"), data = summary_statesB , na.action = na.fail)
r.squaredGLMM(mod1)
mod1 <- glmer(
  ratio5~ stage +sex+ (1|device),family = binomial(link = "logit"), data = summary_statesB , na.action = na.fail)
r.squaredGLMM(mod1)
mod1 <- glmer(
  ratio5~ stage*sex + (1|device),family = binomial(link = "logit"), data = summary_statesB , na.action = na.fail)
r.squaredGLMM(mod1)
mod1 <- glmer(
  ratio5~ (1|device),family = binomial(link = "logit"), data = summary_statesB , na.action = na.fail)
r.squaredGLMM(mod1)
mod1 <- glmer(
  ratio5~ sex +(1|device),family = binomial(link = "logit"), data = summary_statesB , na.action = na.fail)
r.squaredGLMM(mod1)

mod1 <- glmer(
  ratio5~  stage + (1|device),family = binomial(link = "logit"), data = summary_statesB , na.action = na.fail)

mod2 <- glmer(
  ratio5~sex + stage + (1|device),family = binomial(link = "logit"), data = summary_statesB , na.action = na.fail)

confint(mod1, oldNames=FALSE) #acccurate
confint(mod2, oldNames=FALSE) #sex not accurate


mod <- glmer(
  ratio5~  stage + (1|device),family = binomial(link = "logit"), data = summary_statesB , na.action = na.fail)
summary(mod)
AICc(mod) 
confint(mod, oldNames=FALSE) 
r.squaredGLMM(mod)

emm <- emmeans(mod, ~ stage, type = "response")
emm

contrast(emm, method = "pairwise", type = "response")
contrast(emm, method = "revpairwise")

emm.2 <- lsmeans(mod, ~ stage, type = "response") 

summary(emm.2)

ggplot(summary_statesB, aes(x = stage, y = prop_nesting, color = sex)) +
  geom_boxplot()+
  geom_jitter(data = summary_statesB, 
              aes(x = stage, y = prop_nesting, color = sex),
              width = 0.1, alpha = 0.5, size = 0.2) +
  theme_classic()

ggplot(summary_statesB, aes(x = stage, y = prop_nesting)) +
  geom_boxplot()+
  geom_jitter(data = summary_statesB, 
              aes(x = stage, y = prop_nesting),
              width = 0.1, alpha = 0.5, size = 0.2) +
  theme_classic()
```
### Assumption checking

```{r}
mod1 <- mod
resid<-residuals(mod1, type="pearson")

par(mfrow=c(3,2))
# Histogram
hist(resid,col='blue',xlab="residuals",main="")
# Quantile-Quantile plot
qqnorm(resid,pch=16,col='blue',xlab='')
qqline(resid,col='red')

# residuals vs fitted
plot(resid~fitted(mod1)
      , col='blue'
      , pch=16)
abline(h = 0)


boxplot(resid~ summary_statesB$stage, 
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "Stage",
         main = "")
abline(h = 0)
```

### Plots
```{r}

gaa <- ggplot(summary_statesB, aes(x = stage, y = prop_nesting, fill = stage)) +
  geom_boxplot(
    outlier.shape = NA, show.legend = FALSE
  ) +
  geom_point(
    position = position_jitter(width = 0.20),
    size = 0.5,
    alpha = 0.7,
    colour = "black", show.legend = FALSE
  ) +
  scale_fill_manual(values = c("white", "grey60")) +
  scale_y_continuous(limits = c(0,1)) +
  theme_classic()
```

The same process was followed for the other states and for Grand-Lieu